---
title: "Annelida data analyse"
author: "Sam Lambrechts, Io Deflem, Emma Cartuyvels, Hans Van Calster"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())

if (!"phyloseq" %in% rownames(installed.packages())) {
  remotes::install_github("joey711/phyloseq")
}
library(phyloseq)
library(vegan)
library(reshape2)
library(ggplot2)
library(ggpubr)
library(gridExtra)
library(ggbiplot)
library(devtools)
library(factoextra)
library(dplyr)
#### Handig pakket met veel phyloseq add-ons
#install.packages("remotes")
if (!"psadd" %in% rownames(installed.packages())) {
  remotes::install_github("cpauvert/psadd")
}

library(psadd)
library(googledrive)
googledrive::drive_auth(email = "*@inbo.be")

mbag_folder <- "G:/Gedeelde drives/PRJ_MBAG"
```

# Inlezen data

```{r shared-drive-data-folders}
sdf <- shared_drive_find("PRJ_MBAG") %>%
  drive_ls(pattern = "4c_bodem", type = "folder") %>%
  drive_ls(pattern = "data", type = "folder") %>%
  drive_ls(pattern = "statistiek", type = "folder") %>%
  drive_ls(pattern = "Annelida", type = "folder")
```

```{r}
rdata_dribble <- sdf %>%
  drive_ls(pattern = "^phyloseq_Olig01_Annelida$", type = "folder") %>%
  drive_ls("physeq_Olig01_Annelida_species.Rdata") %>%
  drive_reveal(what = "path")
```

```{r}
load(file.path(mbag_folder, rdata_dribble$path))
```


# Alfa-diversiteit

## True Rarefaction

<!--vergeet HIER niet te starten van de NIET-rarefied versie-->

Link: https://jacobrprice.github.io/2017/08/26/phyloseq-to-vegan-and-back.html

```{r}
sample_sums(physeq_Olig01_Annelida_species) ##sanity check

## Pick whatever dataset you want to work with

physeq <- physeq_Olig01_Annelida_species

# physeq <- physeq_18S_Annelida
```

```{r}

##### cleanup ######

physeq <- subset_taxa(physeq, phylum =="Annelida")

sample_sums(physeq)

## Remove samples according to the dataset you want to work with (e.g. without replicates or without a certain land use type)
## metadata staat voorlopig in het nederlands, eerst vertalen naar engels?

# physeq <- subset_samples_no_zero(physeq, Cmon_PlotID != "VUL_IN")


```

```{r}

# Sanity check

physeq

head(sample_data(physeq))

physeq_Olig01_Annelida_species

sample_sums(physeq)


```

```{r}

######################################################################
############# Voorbereidende stap #########################################
######################################################################


# convert the otu_table() within a phyloseq object to a vegan compatible data object
psotu2veg <- function(physeq) {
  OTU <- otu_table(physeq)
  if (taxa_are_rows(OTU)) {
    OTU <- t(OTU)
  }
  return(as(OTU, "matrix"))
}

veganobject <- psotu2veg(physeq)


```

```{r}

###################################################################
############ Perform true rarefaction to estiamte species richness ############
##############################################################################

sam.new <-   data.frame(sample_data(physeq))

# Sanity check

head(sam.new)

sam.new

```

```{r}

########################################################
############### True rarefaction using vegan ###################
########################################################


Estimated_species_richness <- vegan::rarefy(veganobject, 30780, se = FALSE, MARGIN = 1)

Estimated_species_richness # Sanity check

write.table(Estimated_species_richness, file = "true_rarefaction_Estimated_species_richness_Olig01_Annelida.txt")

head(Estimated_species_richness) # Sanity check

true_rarefaction_results <- cbind(sam.new, Estimated_species_richness)

getwd()

write.csv(true_rarefaction_results, "true_rarefaction_results_Olig01_Annelida_final.csv")

```
plot results land use types


```{r}


library(ggplot2)
library(ggrepel)

ggplot(
  true_rarefaction_results,
  aes(x = Landgebruik_MBAG,
      y = Estimated_species_richness,
      color = Landgebruik_MBAG)) +
  geom_point() +
  ylab("Expected species richness (Annelida(Olig01))") +
  xlab("Landgebruik") +
  labs(color = "Landgebruik")


####### Eventueel nog opsmukken met: #################

# namen veranderen naar waarden Landgebruik_MBAG

```

```{r}

###### Diepte #######


ggplot(
  true_rarefaction_results,
  aes(x = Diepte, y = Estimated_species_richness, color = Diepte)) +
  geom_point() +
  ylab("Expected species richness (Annelida(Olig01))") +
  xlab("Diepte") +
  labs(color = "Diepte")

```

```{r}

#####################################################################
########################   Statistics     ###################################
#####################################################################


shapiro.test(true_rarefaction_results$Estimated_species_richness)

# is p-value van de shapiro test hierboven kleiner dan 0.05, dan is de data niet normaal verdeeld en doen we kruskal-wallis
```

```{r}

library(rstatix)

kruskal_test(
  true_rarefaction_results,
  Estimated_species_richness ~ Landgebruik_MBAG
)
kruskal_test(
  true_rarefaction_results,
  Estimated_species_richness ~ Diepte)

```

```{r}




#### alfa-diversiteit 2: Kruskal Wallis en Dunn post-hoc test with adjusted p-values using the Benjamini–Hochberg method

# Eerst beslissen voor hoeveel dingen je wil testen. Welke zijn dus de goede variabelen in de metadata

install.packages("dunn.test")
library(dunn.test)
library(rstatix)


shapiro.test(true_rarefaction_results$Estimated_species_richness)
```

```{r}

# is p-value van de shapiro test hierboven kleiner dan 0.05, dan is de data niet normaal verdeeld en doen we kruskal-wallis


kruskal_test(true_rarefaction_results, Estimated_species_richness ~ Landgebruik_MBAG)
kruskal_test(true_rarefaction_results, Estimated_species_richness ~ Diepte)

```

```{r}

# Define the variable for which you found a significant difference
variable_of_interest <- "Landgebruik_MBAG"  # Replace with the variable name

# Perform the Kruskal-Wallis test
kruskal_result <- kruskal_test(true_rarefaction_results, Estimated_species_richness ~ get(variable_of_interest))


```

```{r}

# Perform the Dunn post-hoc test with Benjamini-Hochberg correction
dunn_result <- dunn.test(
  true_rarefaction_results$Estimated_species_richness,
  true_rarefaction_results[[variable_of_interest]],
  method = "bh")

# Display the Dunn test results
print("Dunn Test Results:")
print(dunn_result)

# Extract adjusted p-values
adjusted_p_values <- dunn_result$P.adjusted

# Print the adjusted p-values for each pairwise comparison
print("Adjusted P-values:")
print(adjusted_p_values)

```

Identify Groups with Highest/Lowest Alpha Diversity: You can also extract specific values if you want to determine which group has the highest or lowest estimated species richness. For example, you can use the summary function to compute summary statistics and identify the group with the maximum or minimum mean estimated species richness:


```{r}

  

# Find the group with the highest estimated species richness
sumstats <- true_rarefaction_results %>%
  as_tibble() %>%
  group_by(Landgebruik_MBAG) %>%
  summarise(
    min = min(Estimated_species_richness),
    max = max(Estimated_species_richness))

knitr::kable(sumstats)
```


```{r rarefaction}






#### rarefaction curves



library(vegan)

S <- specnumber(veganobject)

getwd()

write.table(S, file = "Olig01_Annelida_specnumber.txt")

specnumber(veganobject)

min(rowSums(veganobject))

raremax <- min(rowSums(veganobject))

options(scipen=5)

rarecurve(veganobject, step = 100, sample = raremax, col = "blue", cex= 0.6)



```
General results from phyloseq object

OTU counts per phylum / class / order / family / genus

```{r}



# Load required packages
library(phyloseq)
library(dplyr)

# Count OTUs per Phylum
otu_counts_per_phylum <- tax_table(physeq) %>%
  as.data.frame() %>%
  group_by(Phylum) %>%
  summarise(OTU_Count = n())

# Print the result
print(otu_counts_per_phylum)

write.csv(otu_counts_per_phylum, file = "otu_counts_per_phylum.csv")
```

```{r}

## Genus level

# Load required packages
library(phyloseq)
library(dplyr)

# Count OTUs per Genus
otu_counts_per_genus <- tax_table(physeq) %>%
  as.data.frame() %>%
  group_by(Genus) %>%
  summarise(OTU_Count = n())

# Print the result
print(otu_counts_per_genus)

write.csv(otu_counts_per_genus, file = "otu_counts_per_genus.csv")
```

```{r}

## Species level

# Load required packages
library(phyloseq)
library(dplyr)

# Count OTUs per Species
otu_counts_per_species <- tax_table(physeq) %>%
  as.data.frame() %>%
  group_by(Species) %>%
  summarise(OTU_Count = n())

# Print the result
print(otu_counts_per_species)

write.csv(otu_counts_per_species, file = "otu_counts_per_species.csv")

```

```{r}

##Read counts per phylum   #### werkt miss niet

# Summarize read counts per Phylum
physeq %>%
  tax_glom(taxrank = "Phylum") %>%
  transform_sample_counts(function(x) sum(x)) %>%
  otu_table() %>%
  as.data.frame() %>%
  group_by(Phylum) %>%
  summarise(Read_Count = sum(!!sym(colnames(.))))

# Print the result
print(otu_counts_per_phylum)

```

```{r}
#### How to remove unwanted OTUs from phyloseq object

pop_taxa = function(physeq, badTaxa){
  allTaxa = taxa_names(physeq)
  allTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(allTaxa, physeq))
}

badTaxa = c("Zotu511")
physeq <- pop_taxa(physeq, badTaxa)

physeq


```

```{r}

################################################################
#############   Ordinatie   ####################################



library(phyloseq)
library(vegan)
library(compositions)
library(ggplot2)

# Load phyloseq object
physeq <- physeq_Olig01_Annelida_rar.rarefied.species

physeq

#physeq <- rarefy_even_depth(physeq, sample.size = 30780)

physeq


head(sample_data(physeq))


```

```{r eval=FALSE}
# Load metadata

## OF ZONDER stringsAsFactors = T ???

# Perform CLR transformation
clr_physeq <- transform(physeq, "clr")

# Calculate Bray distance matrix based on the CLR-transformed data
bray_distance <- distance(physeq, method = "bray")

# Perform PCoA
pcoa_result <- ordinate(physeq, method = "PCoA", distance = "bray")

# Create the ordination plot with color based on land_use and shape based on diepte
plot_ordination(physeq, pcoa_result, type = "samples", color = "Landgebruik_MBAG", shape = "Diepte") +
  theme_minimal()





```

