
### `r stringr::str_to_sentence("{{group}} - {{primerset}} - {{unit}}")`

```{r}
group <- "{{group}}"
primerset <- "{{primerset}}"
unit <- "{{unit}}"
season_run <- "{{season_run}}"

```

Seasonal filtering inside the child

```{r}
# Seasonal filtering inside the child
months_map <- list(
  allseasons     = NULL,
  winterhalfjaar = c(11, 12, 1, 2, 3, 4),
  winter         = c(12, 1, 2, 3, 4)
)

data_subset <- combined %>%
  {
    sel <- months_map[[season_run]]
    if (!is.null(sel) && "sampling_month" %in% names(.)) {
      dplyr::filter(., sampling_month %in% sel)
    } else {
      .
    }
  } %>%
  dplyr::filter(
    group     == "{{group}}",
    primerset == "{{primerset}}",
    unit      == "{{unit}}"
  ) %>%
  droplevels()
```

```{r}
cat("season_run =", season_run,
    "| rows =", nrow(data_subset),
    "| months =",
    if ("sampling_month" %in% names(data_subset))
      paste(sort(unique(data_subset$sampling_month)), collapse = ",")
    else "NA",
    "\n")

```


```{r}
data_subset %>%
  group_by(landgebruik, diepte) %>%
  summarize(n_observed_not_missing = sum(!is.na(observed))) %>%
  knitr::kable()
```

```{r}
# remove NAs
data_subset <- data_subset %>%
  filter(!is.na(observed))

diepte_ok <- length(unique(data_subset$diepte)) > 1
```


#### Basismodel

```{r m-richness-{{group}}-{{primerset}}-{{unit}}}
if (diepte_ok) {
  form_richness <- formula(
  observed ~
    log(total_count + 1)
    + landgebruik
    + diepte
    + landgebruik:diepte
    + (1 | cmon_plot_id)
  )
  ziform <- formula(~ 1 + (1 | cmon_plot_id))
} else {
  form_richness <- formula(
  observed ~
    log(total_count + 1)
    + landgebruik
  )
  ziform <- formula(~ 1)
}

cat("Fitting Poisson model\n")
m_pois <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = poisson()
)
cat("Fitting hurdle Poisson model\n")
m_hpois <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_poisson()
)
cat("Fitting negative binomial model\n")
m_nbinom <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = nbinom2()
)
cat("Fitting hurdle negative binomial model\n")
m_hnbinom <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_nbinom2()
)
cat("Fitting generalized Poisson model\n")
m_genpois <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = genpois()
)
cat("Fitting hurdle generalized Poisson model\n")
m_hgenpois <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_genpois()
)

mlist <- list(
  pois = m_pois,
  hpois = m_hpois,
  nbinom = m_nbinom,
  hnbinom = m_hnbinom,
  genpois = m_genpois,
  hgenpois = m_hgenpois
)
```

Het model met de laagste AIC waarde (hoogste AIC gewicht: `AIC_wt`) heeft de beste fit volgens AIC criterium.
Dit betekent niet noodzakelijk dat dit model geen enkel probleem meer heeft met betrekking tot `zero-inflation` (`deflation`) of over- (onder-) dispersie.

Modellen waarvan de AIC niet berekend kan worden, worden verwijderd.
Deze modellen zijn wellicht niet geconvergeerd.

```{r}
converged <- !is.na(map_dbl(mlist, AIC))
cp <- compare_performance(mlist[converged], metrics = c("AIC"), rank = FALSE) |>
  bind_cols(map_dfr(mlist[converged], inflation_dispersion))
kable(cp, digits = 2)
```

```{r}
m_richness <- mlist[[cp$Name[cp$AIC_wt == max(cp$AIC_wt)]]]
```


Het geselecteerde model is:

```{r}
insight::get_call(m_richness)
```

```{r m-richness-checks-{{group}}-{{primerset}}-{{unit}}, error=TRUE}
p <- performance::check_predictions(m_richness)
plot(p) + facet_zoom(xlim = c(0, 10))
p <- performance::check_overdispersion(m_richness)
p
p %>% plot()
p <- performance::check_residuals(m_richness)
p
p %>% plot()
performance::check_collinearity(m_richness)
performance::check_zeroinflation(m_richness)
```


```{r m-richness-summary-{{group}}-{{primerset}}-{{unit}}}
summary(m_richness) |> print(digits = 2)
```

```{r m-richness-anova-{{group}}-{{primerset}}-{{unit}}}
car::Anova(m_richness) |> print(digits = 2)
```

```{r m-richness-marginal-means-per-depth-{{group}}-{{primerset}}-{{unit}}}
# Estimated marginal means per land use (and per depth if applicable)

emm_landuse <- emmeans(
  m_richness,
  specs = "landgebruik",
  by = if (diepte_ok) "diepte" else NULL,
  data = data_subset,
  type = "response"
)

# Save to CSV with automatic naming
write_csv(
  as.data.frame(emm_landuse),
  file = file.path(output_dir, glue("emmeans_landuse_{group}_{primerset}_{unit}.csv"))
)

# print
emm_landuse


```


```{r m-richness-contrasts-per-depth-response-{{group}}-{{primerset}}-{{unit}}}
# Estimated marginal means per land use (and per depth if applicable), on response scale
emm_landuse <- emmeans(
  m_richness,
  specs = "landgebruik",
  by = if (diepte_ok) "diepte" else NULL,
  data = data_subset,
  type = "response"
)

# Pairwise contrasts (auto adjustment by emmeans; keep CIs & p-values)
contrasts_landuse_response <- contrast(
  emm_landuse,
  method = "pairwise",
  infer = c(TRUE, TRUE)
) %>%
  as.data.frame()

# Save to CSV
write_csv(
  contrasts_landuse_response,
  file = file.path(output_dir, glue("contrasts_landuse_response_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_landuse_response)


```


```{r m-richness-contrasts-per-depth-response-bh-{{group}}-{{primerset}}-{{unit}}}
# Summary table
contrasts_landuse_table <- contrasts_landuse_response %>%
  as_tibble() %>%
  dplyr::select(
    dplyr::any_of(c(
      "diepte",        # only present if diepte_ok == TRUE
      "contrast",
      "estimate",      # default column name
      "ratio",         # appears instead of 'estimate' when type = "response"
      "SE",
      "z.ratio",       # or t.ratio, one of them will always exist with infer=TRUE
      "p.value",
      "asymp.LCL",
      "asymp.UCL"
    ))
  )

# Sort
if (diepte_ok) {
  contrasts_landuse_table <- contrasts_landuse_table %>% arrange(diepte, p.value)
} else {
  contrasts_landuse_table <- contrasts_landuse_table %>% arrange(p.value)
}

# Save
write_csv(
  contrasts_landuse_table,
  file = file.path(output_dir, glue("contrasts_landuse_table_response_{group}_{primerset}_{unit}.csv"))
)

# Print
contrasts_landuse_table

```

```{r m-richness-preds-total-count-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("total_count"),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Voorspeld aantal taxa")
```

```{r m-richness-preds-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("landgebruik", "diepte"[diepte_ok]),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Voorspeld aantal taxa")
```


#### Met fysicochemische data

Stapsgewijze modelselectie:

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "c_density",
  "swc_vol",
  "ph_kcl",
  "textuur_som_klei_zand"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset %>%
  mutate(log_c_density = log(c_density)) %>%  # log-transform predictor
  { 
    # Guard: non-positive c_density cannot be logged
    if (any(!is.finite(.$log_c_density))) {
      stop("c_density bevat niet-positieve waarden; log() is niet gedefinieerd. Filter/aanpassen vereist.")
    } else .
  } %>%
  filter(complete.cases(across(all_of(c(variables_no_collinearity, "log_c_density"))))) %>%
  droplevels()
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```


We voegen alle niet collineaire fysicochemische variabelen toe aan het basismodel.
Voor pH gebruiken we een kwadratische effect.
Dit wordt het `full model`.

```{r extend-model-additional-predictors-{{group}}-{{primerset}}-{{unit}}}
form_richness_full <- update(
  form_richness,
  . ~ .
    + swc_vol
    + log_c_density
    + poly(ph_kcl, 2)
    + textuur_som_klei_zand
)

m_richness_full <- update(
  m_richness,
  formula = form_richness_full,
  data = data_subset,
  na.action = na.fail
)
```

We vergelijken de parameterschattingen van het basis en volledig model.

```{r coefficients-se-{{group}}-{{primerset}}-{{unit}}}
base_model_coefs <- broom.mixed::tidy(m_richness)
full_model_coefs <- broom.mixed::tidy(m_richness_full)

coefs_df <- rbind(
  data.frame(Model = "Base", base_model_coefs),
  data.frame(Model = "Full", full_model_coefs)
)
```

```{r plot-coefficient-estimates-{{group}}-{{primerset}}-{{unit}}}
ggplot(
  coefs_df,
  aes(x = term, y = estimate, fill = Model)
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(
    aes(
      ymin = estimate - std.error,
      ymax = estimate + std.error
    ),
    width = 0.2, position = position_dodge(0.9)
  ) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Coefficient Estimates and Standard Errors",
    x = "Predictor",
    y = "Estimate",
    fill = "Model"
  )
```

```{r save-coefficient-estimates-{{group}}-{{primerset}}-{{unit}}}
readr::write_csv(
  coefs_df,
  file = file.path(
    output_dir,
    glue::glue("coefs_base_vs_full_{group}_{primerset}_{unit}.csv")
  )
)
```


Modelselectie met de functie `dredge`.
We beperken het aantal te verkennen parametercombinaties door te forceren dat `total count` `landgebruik` en `diepte` altijd geschat moeten worden ('by design') en maximaal 7 termen in het model zitten.

```{r model-selection-and-comparison-{{group}}-{{primerset}}-{{unit}}, warning=FALSE}
if (diepte_ok) {
  model_set <- dredge(
    m_richness_full,
    m.lim = c(NA, 7),
    fixed = ~
      cond(log(total_count + 1))
    + cond(landgebruik)
    + cond(diepte)
    + (1 | cmon_plot_id)
  )
} else {
  model_set <- dredge(
    m_richness_full,
    m.lim = c(NA, 7),
    fixed = ~
      cond(log(total_count + 1))
    + cond(landgebruik)
    + (1 | cmon_plot_id)
  )
}
```

Per variabele berekenen we de som van modelgewichten over alle modellen en tellen we in hoeveel modellen de variabele voorkwam:

```{r importance-per-predictor-{{group}}-{{primerset}}-{{unit}}}
importance_df <- sw(model_set)
print(importance_df)
```

Visualisatie van modellen die minder dan 4 AICc verschillen van het model met laagste AICc (rij 1):

```{r}
dd <- subset(model_set, delta < 4)

par(mar = c(3, 5, 6, 4))
plot(dd, labAsExpr = TRUE)
```

```{r save-model-selection-{{group}}-{{primerset}}-{{unit}}}

# -- Full selection table
model_set_tbl <- model_set %>%
  as.data.frame() %>%
  rownames_to_column("model_id") %>%
  arrange(AICc)

write_csv(
  model_set_tbl,
  file.path(output_dir, glue("full_model_set_{group}_{primerset}_{unit}.csv"))
)

# -- ΔAICc < 4 subset
dd_obj <- subset(model_set, delta < 4)
dd_tbl <- dd_obj %>%
  as.data.frame() %>%
  rownames_to_column("model_id") %>%
  arrange(AICc)

write_csv(
  dd_tbl,
  file.path(output_dir, glue("model_set_delta_lt4_{group}_{primerset}_{unit}.csv"))
)

# -- 95% cumulative weight subset
cumw_tbl <- model_set_tbl %>%
  mutate(cum_weight = cumsum(weight)) %>%
  filter(cum_weight <= 0.95)

write_csv(
  cumw_tbl,
  file.path(output_dir, glue("model_set_cumw95_{group}_{primerset}_{unit}.csv"))
)

# Variable importance (sum of weights by term)
imp <- MuMIn::importance(model_set)
imp_tbl <- tibble(term = names(imp), importance = as.numeric(imp)) %>%
  arrange(desc(importance))

write_csv(
  imp_tbl,
  file.path(output_dir, glue("model_importance_{group}_{primerset}_{unit}.csv"))
)

# -- (Reproducibility) Save the objects too
saveRDS(model_set, file.path(output_dir, glue("model_set_{group}_{primerset}_{unit}.rds")))
saveRDS(dd_obj,    file.path(output_dir, glue("model_set_delta_lt4_{group}_{primerset}_{unit}.rds")))


```


Dit is de samenvatting voor het model met de laagste AICc:

```{r}
lowest_aic <- get.models(dd, 1)[[1]]
summary(lowest_aic)
```

```{r m-richness-anova-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
car::Anova(lowest_aic) |> print(digits = 2)
```

```{r m-richness-marginal-means-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
# Estimated marginal means per land use (and per depth if applicable), on response scale
emm_landuse_incl_fyschem <- emmeans(
  lowest_aic,
  specs = "landgebruik",
  by    = if (diepte_ok) "diepte" else NULL,
  data  = data_subset,
  type  = "response"
)

# Keep original filename
write_csv(
  as.data.frame(emm_landuse_incl_fyschem),
  file = file.path(output_dir, glue("emmeans_landuse_incl_fyschem_{group}_{primerset}_{unit}.csv"))
)

# Print
emm_landuse_incl_fyschem

```

```{r m-richness-contrasts-per-depth-response-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
# Pairwise contrasts on response scale (keep CIs & p-values), via emmeans
contrasts_landuse_response_incl_fyschem <- contrast(
  emm_landuse_incl_fyschem,
  method = "pairwise",
  infer  = c(TRUE, TRUE)
) %>%
  as.data.frame()

# Write
write_csv(
  contrasts_landuse_response_incl_fyschem,
  file = file.path(output_dir, glue("contrasts_landuse_incl_fyschem_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_landuse_response_incl_fyschem)

```


```{r m-richness-contrasts-incl-fyschem-adj-{{group}}-{{primerset}}-{{unit}}}
# Summary table, aligned with base-model structure
contrasts_landuse_by_depth_incl_fyschem_adj <- contrasts_landuse_response_incl_fyschem %>%
  as_tibble() %>%
  dplyr::select(
    dplyr::any_of(c(
      "diepte",        # only present if diepte_ok == TRUE
      "contrast",
      "estimate",      # link-scale (absent when type = "response")
      "ratio",         # present when type = "response" (IRR)
      "SE",
      "z.ratio",       # or t.ratio
      "p.value",
      "asymp.LCL",
      "asymp.UCL"
    ))
  ) %>%
  { if (diepte_ok) arrange(., diepte, p.value) else arrange(., p.value) }

# Keep original adjusted filename (content is now the summary table without BH)
write_csv(
  contrasts_landuse_by_depth_incl_fyschem_adj,
  file = file.path(output_dir, glue("contrasts_landuse_incl_fyschem_adj_{group}_{primerset}_{unit}.csv"))
)

# Print
contrasts_landuse_by_depth_incl_fyschem_adj

```


```{r error=TRUE}
p <- performance::check_predictions(lowest_aic)
plot(p) + facet_zoom(xlim = c(0, 10))
p <- performance::check_overdispersion(lowest_aic)
p
p %>% plot()
p <- performance::check_residuals(lowest_aic)
p
p %>% plot()
performance::check_collinearity(lowest_aic)
performance::check_zeroinflation(lowest_aic)
```


```{r}
at <- car::Anova(lowest_aic)
efvars <- rownames(at)[order(-at$Chisq)]
interactingvars <- efvars[grepl(":", efvars)]
interactingvars <- stringr::str_split(interactingvars, ":") %>%
  unlist() %>%
  unique()
efvars <- efvars[!efvars %in% interactingvars]
efvars <- stringr::str_replace_all(
  efvars, "^(\\w+\\()?(.+?)(\\))?$", "\\2"
)
efvars <- ifelse(
  efvars == "diepte:landgebruik", "landgebruik:diepte", efvars
)
efvars <- ifelse(
  efvars == "total_count + 1", "total_count", efvars
)
efvars <- ifelse(
  efvars == "ph_kcl, 2", "ph_kcl", efvars
)

p <- vector(mode = "list", length = length(efvars))
p <- set_names(p, efvars)
for (i in efvars) {
  if (grepl(":", i)) {
    conds <- stringr::str_split_1(i, ":")
    nd <- insight::get_datagrid(
      lowest_aic,
      by = conds,
      length = 50
    )
    pl <- marginaleffects::plot_predictions(
      model = lowest_aic,
      by = conds,
      newdata = nd,
      vcov = TRUE,
      type = "response",
      points = 0.1,
      re.form = NA
    )
  } else {
    nd <- insight::get_datagrid(
      lowest_aic,
      by = i,
      length = 50
    )
    pl <- marginaleffects::plot_predictions(
      lowest_aic,
      by = i,
      newdata = nd,
      vcov = TRUE,
      type = "response",
      points = 0.1,
      re.form = NA
    )
  }
  p[[i]] <- pl
}

p
```

global log_c_density effect. This gives a global (averaged) slope across the different land-use categories

```{r log-c-density-lowest_aic-{{group}}-{{primerset}}-{{unit}}}

# Slope for log(c_density) on the latent (log) scale
tr_add <- emtrends(lowest_aic, ~ 1, var = "log_c_density", mode = "latent")
tr_add_sum <- summary(tr_add, infer = TRUE)

z <- qnorm(0.975)

coef_tbl <- as_tibble(tr_add_sum) %>%
  transmute(beta = log_c_density.trend, SE = SE, p.value = p.value)

inc_tbl <- tibble(
  increment = c("×e (e-fold)", "×2 (doubling)", "×10 (tenfold)", "×1.2 (+20%)"),  
  delta     = c(1,               log(2),          log(10),         log(1.2)),
  code      = c("e",             "2x",            "10x",           "1.2x")   # NEW      
)

result <- crossing(coef_tbl, inc_tbl) %>%
  mutate(
    IRR = exp(beta * delta),
    LCL = exp((beta - z * SE) * delta),
    UCL = exp((beta + z * SE) * delta)
  )

result
```


```{r}
# Requires: result (3-row tibble), group, primerset, season_run
# Also assumes you've already created 'summary_all' in the parent.
if (!exists("summary_all")) summary_all <- tibble::tibble()

library(dplyr)
library(tidyr)
library(stringr)

.schema <- c(
  "group","primerset","landgebruik","season_run",
  "beta","SE",
  "IRR_e","LCL_e","UCL_e",
  "IRR_2x","LCL_2x","UCL_2x",
  "IRR_1.2x","LCL_1.2x","UCL_1.2x",   # NEW
  "IRR_10x","LCL_10x","UCL_10x",
  "p.value"
)

# Convert the 3 increments (e, 2x, 10x) into one wide row with matching names
global_row <- {
  vals <- result %>%
    summarise(
      beta    = first(beta),
      SE      = first(SE),
      p.value = first(p.value)
    )

  wide <- result %>%
  dplyr::select(code, IRR, LCL, UCL) %>%
  tidyr::pivot_wider(
    names_from  = code,
    values_from = c(IRR, LCL, UCL),
    names_glue  = "{.value}_{code}"
  )

  bind_cols(vals, wide) %>%
    mutate(
      group       = group,
      primerset   = primerset,
      landgebruik = "global",
      season_run  = season_run
    ) %>%
    relocate(group, primerset, landgebruik, season_run)
}

# Guarantee full schema exists
missing <- setdiff(.schema, names(global_row))
if (length(missing)) global_row[missing] <- NA_real_
global_row <- global_row %>% select(all_of(.schema))

# Append
summary_all <- bind_rows(summary_all, global_row)


```


Include the interaction between log(c_density) and landuse (to get landuse specific slopes)

```{r}
lowest_aic_int <- update(
  lowest_aic,
  formula.  = ~ . + landgebruik * log_c_density,
  data      = data_subset,
  na.action = na.fail)

```

summary of the model that includes the interaction between log(c_density) and landuse

```{r}
summary(lowest_aic_int)

```

Does this model fit better or worse?

```{r}
anova(lowest_aic, lowest_aic_int, test = "Chisq")  # LRT for interaction

```

Per–land-use IRRs for fold changes in carbon density (no subsetting)

```{r log-c-density-lowest-aic-int-perlanduse-{{group}}-{{primerset}}-{{unit}}}
# Slopes per landgebruik w.r.t. log(c_density) on latent scale
tr_lu <- summary(
  emtrends(lowest_aic_int, ~ landgebruik,
           var = "log_c_density", mode = "latent", infer = TRUE)
)

z <- qnorm(0.975)
irr_lu_fc <- as_tibble(tr_lu) %>%
  transmute(
    landgebruik,
    beta = log_c_density.trend,
    SE   = SE,
    # IRRs for fold changes:
    IRR_e   = exp(beta * 1),          LCL_e   = exp((beta - z*SE) * 1),          UCL_e   = exp((beta + z*SE) * 1),
    IRR_2x  = exp(beta * log(2)),     LCL_2x  = exp((beta - z*SE) * log(2)),     UCL_2x  = exp((beta + z*SE) * log(2)),
    IRR_1.2x  = exp(beta * log(1.2)), LCL_1.2x = exp((beta - z*SE) * log(1.2)),  UCL_1.2x = exp((beta + z*SE) * log(1.2)),  # NEW
    IRR_10x = exp(beta * log(10)),    LCL_10x = exp((beta - z*SE) * log(10)),    UCL_10x = exp((beta + z*SE) * log(10)),
    p.value
  )

print(irr_lu_fc, n = Inf, width = Inf)   # <- shows every column/row

```


Append the **per-landuse** effects (`irr_lu_fc`) to `summary_all`

```{r}
# Requires: irr_lu_fc (per landgebruik tibble), group, primerset, season_run
if (!exists("summary_all")) summary_all <- tibble::tibble()

library(dplyr)

.schema <- c(
  "group","primerset","landgebruik","season_run",
  "beta","SE",
  "IRR_e","LCL_e","UCL_e",
  "IRR_2x","LCL_2x","UCL_2x",
  "IRR_1.2x","LCL_1.2x","UCL_1.2x",   # NEW
  "IRR_10x","LCL_10x","UCL_10x",
  "p.value"
)


lu_rows <- irr_lu_fc %>%
  mutate(
    group       = group,
    primerset   = primerset,
    landgebruik = as.character(landgebruik),
    season_run  = season_run
  ) %>%
  relocate(group, primerset, landgebruik, season_run)

# Guarantee full schema exists
missing <- setdiff(.schema, names(lu_rows))
if (length(missing)) lu_rows[missing] <- NA_real_
lu_rows <- lu_rows %>% select(all_of(.schema))

# Append
summary_all <- bind_rows(summary_all, lu_rows)

```


plots per land-use

```{r}
# --- Pull levels & ranges from the fitted model ---
.mf <- model.frame(lowest_aic_int)  # training frame actually used
stopifnot("landgebruik" %in% names(.mf))  # should be there after you added the interaction
lu_lvls <- levels(.mf$landgebruik)

has_diepte <- "diepte" %in% names(.mf) && is.factor(.mf$diepte)
if (has_diepte) {
  dep_lvls      <- levels(.mf$diepte)
  ref_depth_chr <- if (any(grepl("^0-10", dep_lvls))) "0-10" else dep_lvls[1]
  ref_depth     <- factor(ref_depth_chr, levels = dep_lvls)
}

# Range for log_c_density strictly within training data (avoid extrapolation weirdness)
stopifnot("log_c_density" %in% names(.mf))
rng_log <- range(.mf$log_c_density, na.rm = TRUE)
log_seq <- seq(rng_log[1], rng_log[2], length.out = 80)

# --- Build datagrid from the MODEL (auto-fills all other covariates) ---
nd <- insight::get_datagrid(
  lowest_aic_int,
  by     = c("landgebruik", "log_c_density"),
  length = 80
)

# Force the grid to use *all* landgebruik levels & our log grid
nd <- tidyr::crossing(
  landgebruik   = factor(lu_lvls, levels = lu_lvls),
  log_c_density = log_seq
) %>%
  dplyr::left_join(
    # take a single representative row from get_datagrid per landgebruik to carry other covariates
    nd %>%
      dplyr::group_by(landgebruik) %>%
      dplyr::slice(1) %>%
      dplyr::ungroup() %>%
      dplyr::select(-log_c_density),  # we’ll supply our own sequence
    by = "landgebruik"
  ) %>%
  { if (has_diepte) dplyr::mutate(., diepte = ref_depth) else . } %>%
  dplyr::mutate(c_density = exp(log_c_density))

# --- Predictions with CI; fall back to point preds if CI fails ---
pr <- tryCatch(
  marginaleffects::predictions(
    model   = lowest_aic_int,
    newdata = nd,
    type    = "response",
    vcov    = TRUE,
    re.form = NA
  ),
  error = function(e) {
    est <- stats::predict(lowest_aic_int, newdata = nd, type = "response", re.form = NA)
    dplyr::bind_cols(nd, tibble::tibble(estimate = as.numeric(est),
                                        conf.low = NA_real_, conf.high = NA_real_))
  }
)

has_ci <- "conf.low" %in% names(pr) && any(is.finite(pr$conf.low))

# --- Plot ---
p <- ggplot(pr, aes(c_density, estimate))
if (has_ci) p <- p + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15)
p + geom_line(na.rm = TRUE) +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(
    x = "Carbon density",
    y = "Voorspelde rijkdom",
    title = "Voorspelde rijkdom vs carbon density per landgebruik",
    subtitle = paste0(
      "Bij total_count = ",
      if ("total_count" %in% names(.mf)) round(stats::median(.mf$total_count, na.rm = TRUE)) else "—",
      if (has_diepte) paste0(", diepte = ", as.character(ref_depth)) else ""
    )
  ) +
  theme_bw()

```

alternative: plot on the log scale

```{r}
# Reuse .mf, lu_lvls, has_diepte, ref_depth from above

rng_log <- range(.mf$log_c_density, na.rm = TRUE)
log_seq <- seq(rng_log[1], rng_log[2], length.out = 80)

nd_log <- insight::get_datagrid(
  lowest_aic_int,
  by     = c("landgebruik", "log_c_density"),
  length = 80
)

nd_log <- tidyr::crossing(
  landgebruik   = factor(lu_lvls, levels = lu_lvls),
  log_c_density = log_seq
) %>%
  dplyr::left_join(
    nd_log %>%
      dplyr::group_by(landgebruik) %>%
      dplyr::slice(1) %>%
      dplyr::ungroup() %>%
      dplyr::select(-log_c_density),
    by = "landgebruik"
  ) %>%
  { if (has_diepte) dplyr::mutate(., diepte = ref_depth) else . } %>%
  dplyr::mutate(c_density = exp(log_c_density))

pr_log <- tryCatch(
  marginaleffects::predictions(
    model   = lowest_aic_int,
    newdata = nd_log,
    type    = "response",
    vcov    = TRUE,
    re.form = NA
  ),
  error = function(e) {
    est <- stats::predict(lowest_aic_int, newdata = nd_log, type = "response", re.form = NA)
    dplyr::bind_cols(nd_log, tibble::tibble(estimate = as.numeric(est),
                                            conf.low = NA_real_, conf.high = NA_real_))
  }
)

has_ci_log <- "conf.low" %in% names(pr_log) && any(is.finite(pr_log$conf.low))
breaks_nat  <- pretty(range(nd_log$c_density, na.rm = TRUE), n = 5)

p2 <- ggplot(pr_log, aes(log_c_density, estimate))
if (has_ci_log) p2 <- p2 + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15)
p2 + geom_line(na.rm = TRUE) +
  facet_wrap(~ landgebruik, scales = "free_y") +
  scale_x_continuous(
    name   = "Carbon density",
    breaks = log(breaks_nat),
    labels = signif(breaks_nat, 3)
  ) +
  labs(
    y = "Voorspelde rijkdom",
    title = "Voorspelde rijkdom vs log(carbon density) per landgebruik",
    subtitle = paste0(
      "Bij total_count = ",
      if ("total_count" %in% names(.mf)) round(stats::median(.mf$total_count, na.rm = TRUE)) else "—",
      if (has_diepte) paste0(", diepte = ", as.character(ref_depth)) else ""
    )
  ) +
  theme_bw()

```


#### Correlation carbon density

We voegen carbon density toe aan het basismodel.

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "c_density"
)

nr1 <- nrow(data_subset)
cat("De originele dataset heeft", nr1, "rijen.\n")

data_subset <- data_subset %>%
  mutate(log_c_density = log(c_density)) %>%  # log-transform predictor
  { 
    # Guard: non-positive c_density cannot be logged
    if (any(!is.finite(.$log_c_density))) {
      stop("c_density bevat niet-positieve waarden; log() is niet gedefinieerd. Filter/aanpassen vereist.")
    } else .
  } %>%
  filter(complete.cases(across(all_of(c(variables_no_collinearity, "log_c_density"))))) %>%
  droplevels()

nr2 <- nrow(data_subset)
cat("Na verwijderen van rijen met ontbrekende covariaten heeft de dataset", nr2, "rijen.\n")
```


```{r extend-model-carbon-density-{{group}}-{{primerset}}-{{unit}}}
# Voeg log(c_density) toe aan het basismodel
form_richness_full <- update(form_richness, . ~ . + log_c_density)

m_richness_full <- update(
  m_richness,
  formula   = form_richness_full,
  data      = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full)
```

Global effect as IRR for fold changes (×e, ×2, ×10) across different landuse changes

option 1

```{r}

# Slope for log(c_density) on the latent (log) scale
tr_add <- emtrends(m_richness_full, ~ 1, var = "log_c_density", mode = "latent")
tr_add_sum <- summary(tr_add, infer = TRUE)

z <- qnorm(0.975)

coef_tbl <- as_tibble(tr_add_sum) %>%
  transmute(beta = log_c_density.trend, SE = SE, p.value = p.value)

inc_tbl <- tibble(
  increment = c("×e (e-fold)", "×2 (doubling)", "×10 (tenfold)"),
  delta     = c(1, log(2), log(10))
)

result <- crossing(coef_tbl, inc_tbl) %>%
  mutate(
    IRR = exp(beta * delta),
    LCL = exp((beta - z * SE) * delta),
    UCL = exp((beta + z * SE) * delta)
  )

result

```


Plot: predicted richness vs c_density (averaged over LU & depth)

We vary log_c_density (the model term), then back-transform for the x-axis.

```{r}
library(dplyr)
library(tidyr)
library(marginaleffects)
library(ggplot2)

# Range & grid on the natural scale, then log-transform for the model
min_cd <- min(data_subset$c_density[data_subset$c_density > 0], na.rm = TRUE)
max_cd <- max(data_subset$c_density, na.rm = TRUE)
cd_seq  <- seq(min_cd, max_cd, length.out = 80)

grid_list <- list(
  log_c_density = log(cd_seq),
  landgebruik   = levels(data_subset$landgebruik)
)
if ("diepte" %in% names(data_subset)) {
  grid_list$diepte <- levels(droplevels(data_subset$diepte))
}

nd <- tidyr::crossing(!!!grid_list) %>%
  mutate(
    total_count = if ("total_count" %in% names(data_subset))
      median(data_subset$total_count, na.rm = TRUE) else NA_real_
  )

# Proper averaging with SEs over LU & depth at each log_c_density
ap <- marginaleffects::avg_predictions(
  m_richness_full,
  newdata = nd,
  by      = "log_c_density",
  type    = "response",
  re.form = NA,
  vcov    = TRUE
) %>%
  mutate(c_density = exp(log_c_density))

# Plot: x = c_density (natural units), single averaged curve
ggplot(ap, aes(c_density, estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  labs(
    x = "Carbon density",
    y = "Predicted richness",
    title = "Predicted richness vs carbon density (averaged over land-use & depth)"
  ) +
  theme_bw() +
  # Optional: overlay faint raw points
  { 
    resp_var <- all.vars(formula(m_richness_full))[1]
    list(geom_point(data = data_subset,
                    aes(x = c_density, y = .data[[resp_var]]),
                    inherit.aes = FALSE, alpha = 0.12, size = 0.9))
  }


```

alternative: use plot_predictions() on the log scale

```{r}
nd <- insight::get_datagrid(
  m_richness_full,
  by = c("landgebruik", "diepte", "log_c_density"),
  length = 80
) %>%
  mutate(
    total_count = if ("total_count" %in% names(data_subset))
      median(data_subset$total_count, na.rm = TRUE) else NA_real_
  )

p <- marginaleffects::plot_predictions(
  model     = m_richness_full,
  newdata   = nd,
  condition = "log_c_density",   # <- model term
  type      = "response",
  vcov      = TRUE,
  re.form   = NA,
  points    = 0.1
)

# Relabel x ticks in natural units
p + scale_x_continuous(
  name   = "Carbon density",
  labels = function(x) signif(exp(x), 3)
)

```


Interaction model with log(c_density) (to get landuse specific slopes)

```{r}
form_richness_full_int <- update(form_richness, . ~ . + landgebruik * log_c_density)

m_richness_full_int <- update(
  m_richness,
  formula   = form_richness_full_int,
  data      = data_subset,
  na.action = na.fail
)

```

```{r}
summary(m_richness_full_int)

```

```{r}
anova(m_richness_full, m_richness_full_int, test = "Chisq")  # LRT for interaction

```

Per–land-use IRRs for fold changes (no subsetting)

```{r}
# Slopes per landgebruik w.r.t. log(c_density) on latent scale
tr_lu <- summary(
  emtrends(m_richness_full_int, ~ landgebruik,
           var = "log_c_density", mode = "latent", infer = TRUE)
)

z <- qnorm(0.975)
irr_lu_fc <- as_tibble(tr_lu) %>%
  transmute(
    landgebruik,
    beta = log_c_density.trend,
    SE   = SE,
    # IRRs for fold changes:
    IRR_e   = exp(beta * 1),          LCL_e   = exp((beta - z*SE) * 1),          UCL_e   = exp((beta + z*SE) * 1),
    IRR_2x  = exp(beta * log(2)),     LCL_2x  = exp((beta - z*SE) * log(2)),     UCL_2x  = exp((beta + z*SE) * log(2)),
    IRR_10x = exp(beta * log(10)),    LCL_10x = exp((beta - z*SE) * log(10)),    UCL_10x = exp((beta + z*SE) * log(10)),
    p.value
  )

print(irr_lu_fc, n = Inf, width = Inf)   # <- shows every column/row

```

table version

```{r}
knitr::kable(irr_lu_fc, digits = 3,
             caption = "Per–land-use IRRs for e-, 2-, and 10-fold increases in carbon density")

```


plot

Diagnostics for the interaction model

```{r}

mod <- m_richness_full_int

# Predictions vs observed (zoom to data-driven x-range if ggforce available)
p1 <- performance::check_predictions(mod)
xr <- quantile(data_subset$c_density, c(.01, .99), na.rm = TRUE)

if (requireNamespace("ggforce", quietly = TRUE)) {
  library(ggforce)
  print(plot(p1) + ggforce::facet_zoom(xlim = xr))
} else {
  print(plot(p1))
}

# Overdispersion, residuals, collinearity, zero-inflation
od <- performance::check_overdispersion(mod); print(od); plot(od)
res <- performance::check_residuals(mod);    print(res); plot(res)
performance::check_collinearity(mod)
performance::check_zeroinflation(mod)

```

Plot model-based predictions: richness vs c_density by landgebruik

```{r}
# Grid per landgebruik over c_density; fix depth to a reference for clarity
ref_depth <- if ("diepte" %in% names(data_subset) &&
                 any(grepl("0-10", levels(data_subset$diepte)))) "0-10" else levels(data_subset$diepte)[1]

nd_lu <- tidyr::crossing(
  landgebruik = levels(data_subset$landgebruik),
  c_density   = cd_seq
) %>%
  mutate(
    log_c_density = log(c_density),
    total_count   = if ("total_count" %in% names(data_subset))
                      median(data_subset$total_count, na.rm = TRUE) else NA_real_,
    diepte        = if ("diepte" %in% names(data_subset)) ref_depth else factor(NA)
  )

pr <- marginaleffects::predictions(
  model   = m_richness_full_int,
  newdata = nd_lu,
  type    = "response",
  vcov    = TRUE,
  re.form = NA
)

ggplot(pr, aes(c_density, estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(
    x = "Carbon density",
    y = "Voorspelde rijkdom",
    title = "Voorspelde rijkdom vs carbon density per landgebruik",
    subtitle = paste0("Bij total_count = ",
                      round(median(data_subset$total_count, na.rm = TRUE)),
                      if ("diepte" %in% names(data_subset)) paste0(", diepte = ", ref_depth) else "")
  ) +
  theme_bw()


```

alternative: plot on the log scale

```{r}
# --- Plot vs log(c_density), facetted by landgebruik ---

# define log-scale grid
min_cd <- min(data_subset$c_density[data_subset$c_density > 0], na.rm = TRUE)
max_cd <- max(data_subset$c_density, na.rm = TRUE)
log_seq <- seq(log(min_cd), log(max_cd), length.out = 80)

nd_lu_log <- tidyr::crossing(
  landgebruik   = levels(data_subset$landgebruik),
  log_c_density = log_seq
) %>%
  mutate(
    c_density   = exp(log_c_density),  # for pretty axis labels later if needed
    total_count = if ("total_count" %in% names(data_subset))
                    median(data_subset$total_count, na.rm = TRUE) else NA_real_,
    diepte      = if ("diepte" %in% names(data_subset)) {
                    lv <- levels(droplevels(data_subset$diepte))
                    if (any(grepl("0-10", lv))) "0-10" else lv[1]
                  } else factor(NA)
  )

pr_log <- marginaleffects::predictions(
  model   = m_richness_full_int,
  newdata = nd_lu_log,
  type    = "response",
  vcov    = TRUE,
  re.form = NA
)

# choose natural-unit tick marks for readability
breaks_nat <- pretty(c(min_cd, max_cd), n = 5)

ggplot(pr_log, aes(log_c_density, estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ landgebruik, scales = "free_y") +
  scale_x_continuous(
    name   = "Carbon density",
    breaks = log(breaks_nat),
    labels = signif(breaks_nat, 3)
  ) +
  labs(
    y = "Voorspelde rijkdom",
    title = "Voorspelde rijkdom vs log(carbon density) per landgebruik",
    subtitle = paste0("Bij total_count = ",
                      round(median(data_subset$total_count, na.rm = TRUE)),
                      if ("diepte" %in% names(data_subset)) paste0(", diepte = ", nd_lu_log$diepte[1]) else "")
  ) +
  theme_bw()

```


the below is not applicable anymore when using log(c_density)

Per–land-use effect sizes (IRR) for c_density

IRR per IQR, work on the link (latent) scale and exponentiate

```{r eval=FALSE}
library(emmeans)
iqr_cd <- IQR(data_subset$c_density, na.rm = TRUE)

# Get slopes on the link (log) scale
tr <- summary(
  emtrends(mod, ~ landgebruik, var = "c_density", mode = "latent", infer = TRUE)
) # columns include: landgebruik, c_density.trend, SE, df (Inf), p.value

zcrit <- qnorm(0.975)

irr_tbl <- as.data.frame(tr) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = c_density.trend,
    SE    = SE,
    IRR   = exp(beta * iqr_cd),
    LCL   = exp((beta - zcrit * SE) * iqr_cd),
    UCL   = exp((beta + zcrit * SE) * iqr_cd),
    p.value
  )

irr_tbl


```

keep delta inside emtrends(), do this (still on the latent scale) and then exponentiate the trend directly:

```{r eval=FALSE}
tr_delta <- summary(
  emtrends(mod, ~ landgebruik, var = "c_density",
           mode = "latent", delta = iqr_cd, infer = TRUE)
)
irr_tbl <- as.data.frame(tr_delta) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = c_density.trend,  # already scaled by delta
    SE    = SE,               # SE is scaled too
    IRR   = exp(beta),
    LCL   = exp(beta - qnorm(0.975) * SE),
    UCL   = exp(beta + qnorm(0.975) * SE),
    p.value
  )

irr_tbl

```


#### Correlation bulk density

We voegen bulk density toe aan het basismodel.

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "bd"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset[
  complete.cases(data_subset[, variables_no_collinearity]),
]
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```


```{r extend-model-bulk-density-{{group}}-{{primerset}}-{{unit}}}
form_richness_full <- update(
  form_richness,
  . ~ .
    + bd)

m_richness_full <- update(
  m_richness,
  formula = form_richness_full,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full)
```

global bd effect (IRR per IQR), with 95% CI. This gives a global (averaged) slope across the different land-use categories

```{r}
# Global IRR per IQR for bd from the additive model
iqr_cd <- IQR(data_subset$bd, na.rm = TRUE)
tr_add <- summary(emtrends(m_richness_full, ~ 1, var = "bd", mode = "latent", infer = TRUE))
irr_add <- transform(as.data.frame(tr_add),
                     increment = paste0("per IQR (", round(iqr_cd,2), ")"),
                     IRR = exp(bd.trend * iqr_cd),
                     LCL = exp((bd.trend - 1.96*SE) * iqr_cd),
                     UCL = exp((bd.trend + 1.96*SE) * iqr_cd))
irr_add[, c("increment","IRR","LCL","UCL","p.value")]
```

We voegen bulk density en de interactie met landgebruik toe aan het basismodel (to see land-use–specific slopes).

```{r}
form_richness_full_int <- update(
  form_richness,
  . ~ .
    + landgebruik * bd)

m_richness_full_int <- update(
  m_richness,
  formula = form_richness_full_int,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full_int)

```

```{r}
anova(m_richness_full, m_richness_full_int, test = "Chisq")  # LRT for interaction

```


Report per–land-use slopes of bd (IRR per IQR) without subsetting

```{r}
iqr_cd <- IQR(data_subset$bd, na.rm = TRUE)
emtrends(m_richness_full_int, ~ landgebruik, var = "bd",
         delta = iqr_cd, type = "response", infer = TRUE)

```


plot

Diagnostics for the interaction model

```{r}

mod <- m_richness_full_int

# Predictions vs observed (zoom to data-driven x-range if ggforce available)
p1 <- performance::check_predictions(mod)
xr <- quantile(data_subset$bd, c(.01, .99), na.rm = TRUE)

if (requireNamespace("ggforce", quietly = TRUE)) {
  library(ggforce)
  print(plot(p1) + ggforce::facet_zoom(xlim = xr))
} else {
  print(plot(p1))
}

# Overdispersion, residuals, collinearity, zero-inflation
od <- performance::check_overdispersion(mod); print(od); plot(od)
res <- performance::check_residuals(mod);    print(res); plot(res)
performance::check_collinearity(mod)
performance::check_zeroinflation(mod)

```

Model-based predictions: richness vs bd by landgebruik

```{r}
# Build grid over bd within each land-use (keep your “pinning” lines)
by_vars <- c("landgebruik", "bd")
nd <- insight::get_datagrid(mod, by = by_vars, length = 50)

if ("total_count" %in% names(data_subset)) {
  nd$total_count <- stats::median(data_subset$total_count, na.rm = TRUE)
}
ref_depth <- if ("diepte" %in% names(data_subset) &&
                 any(grepl("0-10", levels(data_subset$diepte)))) "0-10" else levels(data_subset$diepte)[1]
nd$diepte <- ref_depth

# Get predictions (response scale), marginal over REs
pr <- marginaleffects::predictions(
  model   = mod,
  newdata = nd,
  type    = "response",
  vcov    = TRUE,
  re.form = NA
)

# Plot: x = bd, facet by land-use
library(ggplot2)
ggplot(pr, aes(x = bd, y = estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(x = "bulk density", y = "Predicted richness",
       title = "Predicted richness vs bulk density by land-use",
       subtitle = paste0("Predictions at total_count = ", round(stats::median(data_subset$total_count, na.rm = TRUE)),
                         ", depth = ", ref_depth)) +
  theme_bw()


```

```{r}
resp_var <- all.vars(formula(mod))[1]
last_plot() +
  geom_point(data = data_subset,
             aes(x = bd, y = .data[[resp_var]]),
             inherit.aes = FALSE, alpha = 0.12, size = 0.9)

```


Per–land-use effect sizes (IRR) for bd

IRR per IQR, work on the link (latent) scale and exponentiate

```{r}
library(emmeans)
iqr_cd <- IQR(data_subset$bd, na.rm = TRUE)

# Get slopes on the link (log) scale
tr <- summary(
  emtrends(mod, ~ landgebruik, var = "bd", mode = "latent", infer = TRUE)
) # columns include: landgebruik, bd.trend, SE, df (Inf), p.value

zcrit <- qnorm(0.975)

irr_tbl <- as.data.frame(tr) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = bd.trend,
    SE    = SE,
    IRR   = exp(beta * iqr_cd),
    LCL   = exp((beta - zcrit * SE) * iqr_cd),
    UCL   = exp((beta + zcrit * SE) * iqr_cd),
    p.value
  )

irr_tbl


```

keep delta inside emtrends(), do this (still on the latent scale) and then exponentiate the trend directly:

```{r}
tr_delta <- summary(
  emtrends(mod, ~ landgebruik, var = "bd",
           mode = "latent", delta = iqr_cd, infer = TRUE)
)
irr_tbl <- as.data.frame(tr_delta) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = bd.trend,  # already scaled by delta
    SE    = SE,               # SE is scaled too
    IRR   = exp(beta),
    LCL   = exp(beta - qnorm(0.975) * SE),
    UCL   = exp(beta + qnorm(0.975) * SE),
    p.value
  )

irr_tbl

```


Center/scale bd (optional) to stabilize estimation and make the intercept interpretable? Not done yet

```{r eval=FALSE}
data_subset$bd_c <- scale(data_subset$bd, center = TRUE, scale = TRUE)
# then use bd_c instead of bd in both above formulas

```


#### Correlation pH

We voegen pH toe aan het basismodel.

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "ph_kcl"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset[
  complete.cases(data_subset[, variables_no_collinearity]),
]
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```


```{r extend-model-ph-{{group}}-{{primerset}}-{{unit}}}
form_richness_full <- update(
  form_richness,
  . ~ .
    + ph_kcl)

m_richness_full <- update(
  m_richness,
  formula = form_richness_full,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full)
```

global ph_kcl effect (IRR per IQR), with 95% CI. This gives a global (averaged) slope across the different land-use categories

```{r}
# Global IRR per IQR for ph_kcl from the additive model
iqr_cd <- IQR(data_subset$ph_kcl, na.rm = TRUE)
tr_add <- summary(emtrends(m_richness_full, ~ 1, var = "ph_kcl", mode = "latent", infer = TRUE))
irr_add <- transform(as.data.frame(tr_add),
                     increment = paste0("per IQR (", round(iqr_cd,2), ")"),
                     IRR = exp(ph_kcl.trend * iqr_cd),
                     LCL = exp((ph_kcl.trend - 1.96*SE) * iqr_cd),
                     UCL = exp((ph_kcl.trend + 1.96*SE) * iqr_cd))
irr_add[, c("increment","IRR","LCL","UCL","p.value")]
```

We voegen pH en de interactie met landgebruik toe aan het basismodel (to see land-use–specific slopes).

```{r}
form_richness_full_int <- update(
  form_richness,
  . ~ .
    + landgebruik * ph_kcl)

m_richness_full_int <- update(
  m_richness,
  formula = form_richness_full_int,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full_int)

```

```{r}
anova(m_richness_full, m_richness_full_int, test = "Chisq")  # LRT for interaction

```


Report per–land-use slopes of ph_kcl (IRR per IQR) without subsetting

```{r}
iqr_cd <- IQR(data_subset$ph_kcl, na.rm = TRUE)
emtrends(m_richness_full_int, ~ landgebruik, var = "ph_kcl",
         delta = iqr_cd, type = "response", infer = TRUE)

```


plot

Diagnostics for the interaction model

```{r}

mod <- m_richness_full_int

# Predictions vs observed (zoom to data-driven x-range if ggforce available)
p1 <- performance::check_predictions(mod)
xr <- quantile(data_subset$ph_kcl, c(.01, .99), na.rm = TRUE)

if (requireNamespace("ggforce", quietly = TRUE)) {
  library(ggforce)
  print(plot(p1) + ggforce::facet_zoom(xlim = xr))
} else {
  print(plot(p1))
}

# Overdispersion, residuals, collinearity, zero-inflation
od <- performance::check_overdispersion(mod); print(od); plot(od)
res <- performance::check_residuals(mod);    print(res); plot(res)
performance::check_collinearity(mod)
performance::check_zeroinflation(mod)

```

Model-based predictions: richness vs ph_kcl by landgebruik

```{r}
# Build grid over ph_kcl within each land-use (keep your “pinning” lines)
by_vars <- c("landgebruik", "ph_kcl")
nd <- insight::get_datagrid(mod, by = by_vars, length = 50)

if ("total_count" %in% names(data_subset)) {
  nd$total_count <- stats::median(data_subset$total_count, na.rm = TRUE)
}
ref_depth <- if ("diepte" %in% names(data_subset) &&
                 any(grepl("0-10", levels(data_subset$diepte)))) "0-10" else levels(data_subset$diepte)[1]
nd$diepte <- ref_depth

# Get predictions (response scale), marginal over REs
pr <- marginaleffects::predictions(
  model   = mod,
  newdata = nd,
  type    = "response",
  vcov    = TRUE,
  re.form = NA
)

# Plot: x = ph_kcl, facet by land-use
library(ggplot2)
ggplot(pr, aes(x = ph_kcl, y = estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(x = "pH", y = "Predicted richness",
       title = "Predicted richness vs pH by land-use",
       subtitle = paste0("Predictions at total_count = ", round(stats::median(data_subset$total_count, na.rm = TRUE)),
                         ", depth = ", ref_depth)) +
  theme_bw()


```

```{r}
resp_var <- all.vars(formula(mod))[1]
last_plot() +
  geom_point(data = data_subset,
             aes(x = ph_kcl, y = .data[[resp_var]]),
             inherit.aes = FALSE, alpha = 0.12, size = 0.9)

```


Per–land-use effect sizes (IRR) for ph_kcl

IRR per IQR, work on the link (latent) scale and exponentiate

```{r}
library(emmeans)
iqr_cd <- IQR(data_subset$ph_kcl, na.rm = TRUE)

# Get slopes on the link (log) scale
tr <- summary(
  emtrends(mod, ~ landgebruik, var = "ph_kcl", mode = "latent", infer = TRUE)
) # columns include: landgebruik, ph_kcl.trend, SE, df (Inf), p.value

zcrit <- qnorm(0.975)

irr_tbl <- as.data.frame(tr) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = ph_kcl.trend,
    SE    = SE,
    IRR   = exp(beta * iqr_cd),
    LCL   = exp((beta - zcrit * SE) * iqr_cd),
    UCL   = exp((beta + zcrit * SE) * iqr_cd),
    p.value
  )

irr_tbl


```

keep delta inside emtrends(), do this (still on the latent scale) and then exponentiate the trend directly:

```{r}
tr_delta <- summary(
  emtrends(mod, ~ landgebruik, var = "ph_kcl",
           mode = "latent", delta = iqr_cd, infer = TRUE)
)
irr_tbl <- as.data.frame(tr_delta) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = ph_kcl.trend,  # already scaled by delta
    SE    = SE,               # SE is scaled too
    IRR   = exp(beta),
    LCL   = exp(beta - qnorm(0.975) * SE),
    UCL   = exp(beta + qnorm(0.975) * SE),
    p.value
  )

irr_tbl

```


Center/scale ph_kcl (optional) to stabilize estimation and make the intercept interpretable? Not done yet

```{r eval=FALSE}
data_subset$ph_kcl_c <- scale(data_subset$ph_kcl, center = TRUE, scale = TRUE)
# then use ph_kcl_c instead of ph_kcl in both above formulas

```


#### Correlation soil texture

We voegen textuur_som_klei_zand toe aan het basismodel.

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "textuur_som_klei_zand"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset[
  complete.cases(data_subset[, variables_no_collinearity]),
]
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```


```{r extend-model-soil-texture-{{group}}-{{primerset}}-{{unit}}}
form_richness_full <- update(
  form_richness,
  . ~ .
    + textuur_som_klei_zand)

m_richness_full <- update(
  m_richness,
  formula = form_richness_full,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full)
```

global textuur_som_klei_zand effect (IRR per IQR), with 95% CI. This gives a global (averaged) slope across the different land-use categories

```{r}
# Global IRR per IQR for textuur_som_klei_zand from the additive model
iqr_cd <- IQR(data_subset$textuur_som_klei_zand, na.rm = TRUE)
tr_add <- summary(emtrends(m_richness_full, ~ 1, var = "textuur_som_klei_zand", mode = "latent", infer = TRUE))
irr_add <- transform(as.data.frame(tr_add),
                     increment = paste0("per IQR (", round(iqr_cd,2), ")"),
                     IRR = exp(textuur_som_klei_zand.trend * iqr_cd),
                     LCL = exp((textuur_som_klei_zand.trend - 1.96*SE) * iqr_cd),
                     UCL = exp((textuur_som_klei_zand.trend + 1.96*SE) * iqr_cd))
irr_add[, c("increment","IRR","LCL","UCL","p.value")]
```

We voegen textuur_som_klei_zand en de interactie met landgebruik toe aan het basismodel (to see land-use–specific slopes).

```{r}
form_richness_full_int <- update(
  form_richness,
  . ~ .
    + landgebruik * textuur_som_klei_zand)

m_richness_full_int <- update(
  m_richness,
  formula = form_richness_full_int,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full_int)

```

```{r}
anova(m_richness_full, m_richness_full_int, test = "Chisq")  # LRT for interaction

```


Report per–land-use slopes of textuur_som_klei_zand (IRR per IQR) without subsetting

```{r}
iqr_cd <- IQR(data_subset$textuur_som_klei_zand, na.rm = TRUE)
emtrends(m_richness_full_int, ~ landgebruik, var = "textuur_som_klei_zand",
         delta = iqr_cd, type = "response", infer = TRUE)

```


plot

Diagnostics for the interaction model

```{r}

mod <- m_richness_full_int

# Predictions vs observed (zoom to data-driven x-range if ggforce available)
p1 <- performance::check_predictions(mod)
xr <- quantile(data_subset$textuur_som_klei_zand, c(.01, .99), na.rm = TRUE)

if (requireNamespace("ggforce", quietly = TRUE)) {
  library(ggforce)
  print(plot(p1) + ggforce::facet_zoom(xlim = xr))
} else {
  print(plot(p1))
}

# Overdispersion, residuals, collinearity, zero-inflation
od <- performance::check_overdispersion(mod); print(od); plot(od)
res <- performance::check_residuals(mod);    print(res); plot(res)
performance::check_collinearity(mod)
performance::check_zeroinflation(mod)

```

Model-based predictions: richness vs textuur_som_klei_zand by landgebruik

```{r}
# Build grid over textuur_som_klei_zand within each land-use (keep your “pinning” lines)
by_vars <- c("landgebruik", "textuur_som_klei_zand")
nd <- insight::get_datagrid(mod, by = by_vars, length = 50)

if ("total_count" %in% names(data_subset)) {
  nd$total_count <- stats::median(data_subset$total_count, na.rm = TRUE)
}
ref_depth <- if ("diepte" %in% names(data_subset) &&
                 any(grepl("0-10", levels(data_subset$diepte)))) "0-10" else levels(data_subset$diepte)[1]
nd$diepte <- ref_depth

# Get predictions (response scale), marginal over REs
pr <- marginaleffects::predictions(
  model   = mod,
  newdata = nd,
  type    = "response",
  vcov    = TRUE,
  re.form = NA
)

# Plot: x = textuur_som_klei_zand, facet by land-use
library(ggplot2)
ggplot(pr, aes(x = textuur_som_klei_zand, y = estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(x = "textuur_som_klei_zand", y = "Predicted richness",
       title = "Predicted richness vs textuur_som_klei_zand by land-use",
       subtitle = paste0("Predictions at total_count = ", round(stats::median(data_subset$total_count, na.rm = TRUE)),
                         ", depth = ", ref_depth)) +
  theme_bw()


```

```{r}
resp_var <- all.vars(formula(mod))[1]
last_plot() +
  geom_point(data = data_subset,
             aes(x = textuur_som_klei_zand, y = .data[[resp_var]]),
             inherit.aes = FALSE, alpha = 0.12, size = 0.9)

```


Per–land-use effect sizes (IRR) for textuur_som_klei_zand

IRR per IQR, work on the link (latent) scale and exponentiate

```{r}
library(emmeans)
iqr_cd <- IQR(data_subset$textuur_som_klei_zand, na.rm = TRUE)

# Get slopes on the link (log) scale
tr <- summary(
  emtrends(mod, ~ landgebruik, var = "textuur_som_klei_zand", mode = "latent", infer = TRUE)
) # columns include: landgebruik, textuur_som_klei_zand.trend, SE, df (Inf), p.value

zcrit <- qnorm(0.975)

irr_tbl <- as.data.frame(tr) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = textuur_som_klei_zand.trend,
    SE    = SE,
    IRR   = exp(beta * iqr_cd),
    LCL   = exp((beta - zcrit * SE) * iqr_cd),
    UCL   = exp((beta + zcrit * SE) * iqr_cd),
    p.value
  )

irr_tbl


```

keep delta inside emtrends(), do this (still on the latent scale) and then exponentiate the trend directly:

```{r}
tr_delta <- summary(
  emtrends(mod, ~ landgebruik, var = "textuur_som_klei_zand",
           mode = "latent", delta = iqr_cd, infer = TRUE)
)
irr_tbl <- as.data.frame(tr_delta) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = textuur_som_klei_zand.trend,  # already scaled by delta
    SE    = SE,               # SE is scaled too
    IRR   = exp(beta),
    LCL   = exp(beta - qnorm(0.975) * SE),
    UCL   = exp(beta + qnorm(0.975) * SE),
    p.value
  )

irr_tbl

```


Center/scale textuur_som_klei_zand (optional) to stabilize estimation and make the intercept interpretable? Not done yet

```{r eval=FALSE}
data_subset$textuur_som_klei_zand_c <- scale(data_subset$textuur_som_klei_zand, center = TRUE, scale = TRUE)
# then use textuur_som_klei_zand_c instead of textuur_som_klei_zand in both above formulas

```


#### Correlation soil water content

We voegen swc_vol toe aan het basismodel.

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "swc_vol"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset[
  complete.cases(data_subset[, variables_no_collinearity]),
]
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```


```{r extend-model-swc-vol-{{group}}-{{primerset}}-{{unit}}}
form_richness_full <- update(
  form_richness,
  . ~ .
    + swc_vol)

m_richness_full <- update(
  m_richness,
  formula = form_richness_full,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full)
```

global swc_vol effect (IRR per IQR), with 95% CI. This gives a global (averaged) slope across the different land-use categories

```{r}
# Global IRR per IQR for swc_vol from the additive model
iqr_cd <- IQR(data_subset$swc_vol, na.rm = TRUE)
tr_add <- summary(emtrends(m_richness_full, ~ 1, var = "swc_vol", mode = "latent", infer = TRUE))
irr_add <- transform(as.data.frame(tr_add),
                     increment = paste0("per IQR (", round(iqr_cd,2), ")"),
                     IRR = exp(swc_vol.trend * iqr_cd),
                     LCL = exp((swc_vol.trend - 1.96*SE) * iqr_cd),
                     UCL = exp((swc_vol.trend + 1.96*SE) * iqr_cd))
irr_add[, c("increment","IRR","LCL","UCL","p.value")]
```

We voegen swc_vol en de interactie met landgebruik toe aan het basismodel (to see land-use–specific slopes).

```{r}
form_richness_full_int <- update(
  form_richness,
  . ~ .
    + landgebruik * swc_vol)

m_richness_full_int <- update(
  m_richness,
  formula = form_richness_full_int,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full_int)

```

```{r}
anova(m_richness_full, m_richness_full_int, test = "Chisq")  # LRT for interaction

```


Report per–land-use slopes of swc_vol (IRR per IQR) without subsetting

```{r}
iqr_cd <- IQR(data_subset$swc_vol, na.rm = TRUE)
emtrends(m_richness_full_int, ~ landgebruik, var = "swc_vol",
         delta = iqr_cd, type = "response", infer = TRUE)

```


plot

Diagnostics for the interaction model

```{r}

mod <- m_richness_full_int

# Predictions vs observed (zoom to data-driven x-range if ggforce available)
p1 <- performance::check_predictions(mod)
xr <- quantile(data_subset$swc_vol, c(.01, .99), na.rm = TRUE)

if (requireNamespace("ggforce", quietly = TRUE)) {
  library(ggforce)
  print(plot(p1) + ggforce::facet_zoom(xlim = xr))
} else {
  print(plot(p1))
}

# Overdispersion, residuals, collinearity, zero-inflation
od <- performance::check_overdispersion(mod); print(od); plot(od)
res <- performance::check_residuals(mod);    print(res); plot(res)
performance::check_collinearity(mod)
performance::check_zeroinflation(mod)

```

Model-based predictions: richness vs swc_vol by landgebruik

```{r}
# Build grid over swc_vol within each land-use (keep your “pinning” lines)
by_vars <- c("landgebruik", "swc_vol")
nd <- insight::get_datagrid(mod, by = by_vars, length = 50)

if ("total_count" %in% names(data_subset)) {
  nd$total_count <- stats::median(data_subset$total_count, na.rm = TRUE)
}
ref_depth <- if ("diepte" %in% names(data_subset) &&
                 any(grepl("0-10", levels(data_subset$diepte)))) "0-10" else levels(data_subset$diepte)[1]
nd$diepte <- ref_depth

# Get predictions (response scale), marginal over REs
pr <- marginaleffects::predictions(
  model   = mod,
  newdata = nd,
  type    = "response",
  vcov    = TRUE,
  re.form = NA
)

# Plot: x = swc_vol, facet by land-use
library(ggplot2)
ggplot(pr, aes(x = swc_vol, y = estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(x = "swc_vol", y = "Predicted richness",
       title = "Predicted richness vs swc_vol by land-use",
       subtitle = paste0("Predictions at total_count = ", round(stats::median(data_subset$total_count, na.rm = TRUE)),
                         ", depth = ", ref_depth)) +
  theme_bw()


```

```{r}
resp_var <- all.vars(formula(mod))[1]
last_plot() +
  geom_point(data = data_subset,
             aes(x = swc_vol, y = .data[[resp_var]]),
             inherit.aes = FALSE, alpha = 0.12, size = 0.9)

```


Per–land-use effect sizes (IRR) for swc_vol

IRR per IQR, work on the link (latent) scale and exponentiate

```{r}
library(emmeans)
iqr_cd <- IQR(data_subset$swc_vol, na.rm = TRUE)

# Get slopes on the link (log) scale
tr <- summary(
  emtrends(mod, ~ landgebruik, var = "swc_vol", mode = "latent", infer = TRUE)
) # columns include: landgebruik, swc_vol.trend, SE, df (Inf), p.value

zcrit <- qnorm(0.975)

irr_tbl <- as.data.frame(tr) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = swc_vol.trend,
    SE    = SE,
    IRR   = exp(beta * iqr_cd),
    LCL   = exp((beta - zcrit * SE) * iqr_cd),
    UCL   = exp((beta + zcrit * SE) * iqr_cd),
    p.value
  )

irr_tbl


```

keep delta inside emtrends(), do this (still on the latent scale) and then exponentiate the trend directly:

```{r}
tr_delta <- summary(
  emtrends(mod, ~ landgebruik, var = "swc_vol",
           mode = "latent", delta = iqr_cd, infer = TRUE)
)
irr_tbl <- as.data.frame(tr_delta) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = swc_vol.trend,  # already scaled by delta
    SE    = SE,               # SE is scaled too
    IRR   = exp(beta),
    LCL   = exp(beta - qnorm(0.975) * SE),
    UCL   = exp(beta + qnorm(0.975) * SE),
    p.value
  )

irr_tbl

```


Center/scale swc_vol (optional) to stabilize estimation and make the intercept interpretable? Not done yet

```{r eval=FALSE}
data_subset$swc_vol_c <- scale(data_subset$swc_vol, center = TRUE, scale = TRUE)
# then use swc_vol_c instead of swc_vol in both above formulas

```




