
### `r stringr::str_to_sentence("{{group}} - {{primerset}} - {{unit}}")`

```{r}
group <- "{{group}}"
primerset <- "{{primerset}}"
unit <- "{{unit}}"
```

```{r}
data_subset <- combined %>%
  filter(
    group == "{{group}}",
    primerset == "{{primerset}}",
    unit == "{{unit}}"
  )
```


```{r}
data_subset %>%
  group_by(landgebruik, diepte) %>%
  summarize(n_observed_not_missing = sum(!is.na(observed))) %>%
  knitr::kable()
```

```{r}
# remove NAs
data_subset <- data_subset %>%
  filter(!is.na(observed))

diepte_ok <- length(unique(data_subset$diepte)) > 1
```


#### Basismodel

```{r m-richness-{{group}}-{{primerset}}-{{unit}}}
if (diepte_ok) {
  form_richness <- formula(
  observed ~
    log(total_count + 1)
    + landgebruik
    + diepte
    + landgebruik:diepte
    + (1 | cmon_plot_id)
  )
} else {
  form_richness <- formula(
  observed ~
    log(total_count + 1)
    + landgebruik
    + (1 | cmon_plot_id)
  )
}

ziform <- formula(~ 1 + (1 | cmon_plot_id))

cat("Fitting Poisson model\n")
m_pois <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = poisson()
)
cat("Fitting hurdle Poisson model\n")
m_hpois <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_poisson()
)
cat("Fitting negative binomial model\n")
m_nbinom <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = nbinom2()
)
cat("Fitting hurdle negative binomial model\n")
m_hnbinom <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_nbinom2()
)
cat("Fitting generalized Poisson model\n")
m_genpois <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = genpois()
)
cat("Fitting hurdle generalized Poisson model\n")
m_hgenpois <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_genpois()
)

mlist <- list(
  pois = m_pois,
  hpois = m_hpois,
  nbinom = m_nbinom,
  hnbinom = m_hnbinom,
  genpois = m_genpois,
  hgenpois = m_hgenpois
)
```

Het model met de laagste AIC waarde (hoogste AIC gewicht: `AIC_wt`) heeft de beste fit volgens AIC criterium.
Dit betekent niet noodzakelijk dat dit model geen enkel probleem meer heeft met betrekking tot `zero-inflation` (`deflation`) of over- (onder-) dispersie.

Modellen waarvan de AIC niet berekend kan worden, worden verwijderd.
Deze modellen zijn wellicht niet geconvergeerd.

```{r}
converged <- !is.na(map_dbl(mlist, AIC))
cp <- compare_performance(mlist[converged], metrics = c("AIC"), rank = FALSE) |>
  bind_cols(map_dfr(mlist[converged], inflation_dispersion))
kable(cp, digits = 2)
```

```{r}
m_richness <- mlist[[cp$Name[cp$AIC_wt == max(cp$AIC_wt)]]]
```


Het geselecteerde model is:

```{r}
insight::get_call(m_richness)
```

```{r m-richness-checks-{{group}}-{{primerset}}-{{unit}}, error=TRUE}
p <- performance::check_predictions(m_richness)
plot(p) + facet_zoom(xlim = c(0, 10))
p <- performance::check_overdispersion(m_richness)
p
p %>% plot()
p <- performance::check_residuals(m_richness)
p
p %>% plot()
performance::check_collinearity(m_richness)
performance::check_zeroinflation(m_richness)
```


```{r m-richness-summary-{{group}}-{{primerset}}-{{unit}}}
summary(m_richness) |> print(digits = 2)
```

```{r m-richness-anova-{{group}}-{{primerset}}-{{unit}}}
car::Anova(m_richness) |> print(digits = 2)
```

```{r m-richness-marginal-means-per-depth-{{group}}-{{primerset}}-{{unit}}}
# Estimated marginal means per land use *within each depth layer*
emmeans_landuse_by_depth <- marginaleffects::avg_predictions(
  m_richness,
  variables = "landgebruik",
  newdata = data_subset,
  by = c("diepte", "landgebruik"),
  type = "response"
)

# Save to CSV with automatic naming
write_csv(
  emmeans_landuse_by_depth,
  file = file.path(output_dir, glue("emmeans_landuse_by_depth_{group}_{primerset}_{unit}.csv"))
)

# print
emmeans_landuse_by_depth
```

```{r m-richness-marginaleffects-contrasts-{{group}}-{{primerset}}-{{unit}}}
# Pairwise contrasts between land use types *within each depth layer*
contrasts_landuse_by_depth <- marginaleffects::comparisons(
  model = m_richness,
  variables = "landgebruik",
  newdata = data_subset,
  type = "response",
  by = "diepte")

# Save to CSV with automatic naming
write_csv(
  contrasts_landuse_by_depth,
  file = file.path(output_dir, glue("marginaleffects_contrasts_landuse_by_depth_{group}_{primerset}_{unit}.csv"))
)

# print
contrasts_landuse_by_depth
```

```{r m-richness-emmeans-contrasts-per-depth-{{group}}-{{primerset}}-{{unit}}}
# Estimate marginal means per land use per depth
emm_landuse_by_depth <- emmeans(
  m_richness,
  specs = "landgebruik",
  by = "diepte",
  data = data_subset
)

# Compute pairwise contrasts
contrasts_landuse_by_depth <- contrast(emm_landuse_by_depth, method = "pairwise", adjust = "none") %>%
  as.data.frame()

# Save to CSV
write_csv(
  contrasts_landuse_by_depth,
  file = file.path(output_dir, glue("contrasts_landuse_by_depth_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_landuse_by_depth)

```

```{r m-richness-contrasts-per-depth-bh-{{group}}-{{primerset}}-{{unit}}}
# Add BH-adjusted p-values per depth
contrasts_landuse_by_depth_corrected <- contrasts_landuse_by_depth %>%
  as_tibble() %>%
  group_by(diepte) %>%
  mutate(p_bh = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  select(
    diepte,
    contrast,
    estimate,
    SE,
    z.ratio,
    p.value,
    p_bh
  ) %>%
  arrange(diepte, p.value)


# Save corrected results to CSV
write_csv(
  contrasts_landuse_by_depth_corrected,
  file = file.path(output_dir, glue("contrasts_landuse_by_depth_bh_{group}_{primerset}_{unit}.csv"))
)

# Print
contrasts_landuse_by_depth_corrected

```

```{r m-richness-contrasts-per-depth-response-{{group}}-{{primerset}}-{{unit}}}
# Estimate marginal means per land use per depth
emm_landuse_by_depth <- emmeans(
  m_richness,
  specs = "landgebruik",
  by = "diepte",
  data = data_subset,
  type = "response"
)

# Compute pairwise contrasts
contrasts_landuse_by_depth_response <- contrast(emm_landuse_by_depth, method = "pairwise", adjust = "none", infer = c(TRUE, TRUE)) %>%
  as.data.frame()

# Save to CSV
write_csv(
  contrasts_landuse_by_depth_response,
  file = file.path(output_dir, glue("contrasts_landuse_by_depth_response_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_landuse_by_depth_response)

```


```{r m-richness-contrasts-per-depth-response-bh-{{group}}-{{primerset}}-{{unit}}}
# Add BH-adjusted p-values per depth
contrasts_landuse_by_depth_corrected <- contrasts_landuse_by_depth_response %>%
  as_tibble() %>%
  group_by(diepte) %>%
  mutate(p_bh = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  select(
    diepte,
    contrast,
    ratio,
    SE,
    z.ratio,
    p.value,
    p_bh,
    asymp.LCL,
    asymp.UCL
  ) %>%
  arrange(diepte, p.value)


# Save corrected results to CSV
write_csv(
  contrasts_landuse_by_depth_corrected,
  file = file.path(output_dir, glue("contrasts_landuse_by_depth_bh_response_{group}_{primerset}_{unit}.csv"))
)

# Print
contrasts_landuse_by_depth_corrected

```

```{r m-richness-preds-values-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
predicted_values <- marginaleffects::avg_predictions(
  m_richness,
  variables = c("landgebruik", "diepte"),
  newdata = data_subset,
  type = "response"
)

# save to CSV
write_csv(
  predicted_values,
  file = file.path(output_dir, glue("predicted_values_{group}_{primerset}_{unit}.csv"))
)

# View
print(predicted_values)
```


```{r m-richness-preds-total-count-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("total_count"),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Voorspeld aantal taxa")
```

```{r m-richness-preds-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("landgebruik", "diepte"[diepte_ok]),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Voorspeld aantal taxa")
```


#### Met fysicochemische data

##### Stapsgewijze modelselectie:

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "c_density",
  "cn_stockbased",
  "swc_vol",
  "ph_kcl",
  "bd",
  "textuur_som_klei_zand"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset[
  complete.cases(data_subset[, variables_no_collinearity]),
]
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```


We voegen alle niet collineaire fysicochemische variabelen toe aan het basismodel.
Voor pH gebruiken we een kwadratische effect.
Dit wordt het `full model`.

```{r extend-model-additional-predictors-{{group}}-{{primerset}}-{{unit}}}
form_richness_full <- update(
  form_richness,
  . ~ .
    + swc_vol
    + c_density
    + cn_stockbased
    + poly(ph_kcl, 2)
    + bd
    + textuur_som_klei_zand
)

m_richness_full <- update(
  m_richness,
  formula = form_richness_full,
  data = data_subset,
  na.action = na.fail
)
```

We vergelijken de parameterschattingen van het basis en volledig model.

```{r coefficients-se-{{group}}-{{primerset}}-{{unit}}}
# Compare parameter estimates (fixed effects in the conditional model only)
# coefficients-se-{{group}}-{{primerset}}-{{unit}}
base_model_coefs <- broom.mixed::tidy(m_richness)
full_model_coefs <- broom.mixed::tidy(m_richness_full)

coefs_df <- dplyr::bind_rows(
  dplyr::mutate(base_model_coefs, Model = "Base"),
  dplyr::mutate(full_model_coefs, Model = "Full")
) |>
  # Keep only fixed effects from the conditional (count) part
  dplyr::filter(effect == "fixed", component == "cond") |>
  # Drop the intercept
  dplyr::filter(term != "(Intercept)") |>
  # Remove rows without SE (just in case) and order by |estimate|
  dplyr::filter(!is.na(std.error)) |>
  dplyr::mutate(term = forcats::fct_reorder(term, abs(estimate), .fun = max))
```

```{r plot-coefficient-estimates-{{group}}-{{primerset}}-{{unit}}}
# plot-coefficient-estimates-{{group}}-{{primerset}}-{{unit}}
ggplot(coefs_df, aes(x = term, y = estimate, color = Model)) +
  geom_pointrange(aes(ymin = estimate - std.error, ymax = estimate + std.error),
                  position = position_dodge(width = 0.5)) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Conditional model (fixed effects only): estimates ± SE",
    x = "Predictor",
    y = "Estimate"
  )
```


Modelselectie met de functie `dredge`.
We beperken het aantal te verkennen parametercombinaties door te forceren dat `total count` `landgebruik` en `diepte` altijd geschat moeten worden ('by design') en maximaal 7 termen in het model zitten.

```{r model-selection-and-comparison-{{group}}-{{primerset}}-{{unit}}, warning=FALSE}
if (diepte_ok) {
  model_set <- dredge(
    m_richness_full,
    m.lim = c(NA, 7),
    fixed = ~
      cond(log(total_count + 1))
    + cond(landgebruik)
    + cond(diepte)
    + (1 | cmon_plot_id)
  )
} else {
  model_set <- dredge(
    m_richness_full,
    m.lim = c(NA, 7),
    fixed = ~
      cond(log(total_count + 1))
    + cond(landgebruik)
    + (1 | cmon_plot_id)
  )
}
```

##### term importance across the supported model set

Per variabele berekenen we de som van modelgewichten over alle modellen en tellen we in hoeveel modellen de variabele voorkwam:

```{r importance-per-predictor-{{group}}-{{primerset}}-{{unit}}}
importance_df <- sw(model_set)
print(importance_df)
```

Visualisatie van modellen die minder dan 4 AICc verschillen van het model met laagste AICc (rij 1):

```{r}
dd <- subset(model_set, delta < 4)

par(mar = c(3, 5, 6, 4))
plot(dd, labAsExpr = TRUE)
```

check the cumulative weight that dd captures

```{r}

ms_all  <- as.data.frame(model_set)
ms_top  <- as.data.frame(dd)

# match by row names (model indices)
covered <- sum(ms_all$weight[ rownames(ms_all) %in% rownames(ms_top) ], na.rm=TRUE)
covered
# e.g. 0.83 means your ΔAICc ≤ 4 set covers 83% of the total weight


```

```{r}
# 2) Sanity checks
nrow(as.data.frame(dd))          # number of models in the top set
nrow(as.data.frame(model_set))   # total number of dredged models
sum(as.data.frame(model_set)$weight, na.rm = TRUE)  # should be ~1
```

```{r}
# 3) (Optional) Find the smallest set reaching 95% cumulative weight
ms_all <- as.data.frame(model_set)[order(model_set$delta), ]
ms_all$cum_w <- cumsum(ms_all$weight)
cut95 <- which(ms_all$cum_w >= 0.95)[1]
list(
  n_models_95 = cut95,
  cum_weight_95 = ms_all$cum_w[cut95],
  delta_at_cut = ms_all$delta[cut95]
)
```


Get summed Akaike weights per term (workaround for handling glmmTMB/hurdle quirks)

```{r}
ms <- as.data.frame(dd)

term_cols <- setdiff(names(ms), c("df","logLik","AICc","delta","weight"))
sum_wt <- sapply(term_cols, \(cl) sum(ms$weight[ms[[cl]] == 1], na.rm = TRUE))
nmods  <- sapply(term_cols, \(cl) sum(ms[[cl]] == 1, na.rm = TRUE))

importance_tbl <- data.frame(term = term_cols,
                             sum_weight = sum_wt,
                             n_models   = nmods) |>
  dplyr::arrange(dplyr::desc(sum_weight))
importance_tbl

```

same but group polynomial pH terms so you get one importance value for pH

```{r}

# 3) Columns to ignore (meta + fixed-by-design + structural intercepts)
fixed_terms <- if (diepte_ok) {
  c("cond(log(total_count + 1))", "cond(landgebruik)", "cond(diepte)")
} else {
  c("cond(log(total_count + 1))", "cond(landgebruik)")
}
drop_cols <- c("df","logLik","AICc","delta","weight",
               fixed_terms,
               "cond((Int))","zi((Int))","disp((Int))")

# 4) Candidate term columns (0/1 indicators)
term_cols <- setdiff(names(ms), drop_cols)

# 5) Group polynomial pH terms so they are treated as ONE predictor
#    dredge typically names them like "cond(poly(ph_kcl, 2))1" and "...2"
group_name <- function(x) gsub("\\)\\d+$", ")", x)  # collapse ")1"/")2" to ")"
groups <- unique(group_name(term_cols))

# 6) Summed weights and model counts per (grouped) term
importance_tbl <- lapply(groups, function(g) {
  cols <- term_cols[group_name(term_cols) == g]
  # A model "has" the grouped term if ANY of its columns are 1
  has_term <- rowSums(ms[ , cols, drop = FALSE], na.rm = TRUE) > 0
  data.frame(
    term        = g,
    sum_weight  = sum(ms$weight[has_term], na.rm = TRUE),
    n_models    = sum(has_term, na.rm = TRUE)
  )
}) %>% bind_rows() %>%
  arrange(desc(sum_weight))

# 7) (Optional) Add the fixed terms with weight 1.0 for clarity
if (length(fixed_terms)) {
  fixed_tbl <- data.frame(
    term = fixed_terms,
    sum_weight = 1.0,
    n_models = nrow(ms)
  )
  importance_tbl <- bind_rows(fixed_tbl, importance_tbl) %>%
    arrange(desc(sum_weight))
}

importance_tbl

```

##### effect size: model-averaged effects 

(MuMIn::model.avg) and plotted on the response scale.

```{r}
library(MuMIn)
top_models <- get.models(dd, 1:nrow(dd))
ma <- model.avg(top_models)  # conditional averaging

sum_ma <- summary(ma)
tab <- as.data.frame(sum_ma$coefmat.subset)  # CONDITIONAL averages
tab$term <- rownames(tab)

# keep fixed effects from conditional part; drop intercept
cond_tab <- tab %>%
  filter(str_starts(term, "cond("), term != "cond((Int))") %>%
  transmute(
    term_raw = term,
    term     = str_remove_all(term, "^cond\\(|\\)$"),
    estimate = Estimate,
    se       = `Std. Error`,
    lwr      = estimate - 1.96*se,
    upr      = estimate + 1.96*se
  )

```

Continuous main effects → % change per IQR

```{r}
# helper to extract base variable name
base_var <- function(s) {
  if (grepl("^poly\\(", s)) sub("^poly\\(([^,]+),.*$", "\\1", s)
  else if (grepl(":", s)) NA_character_               # interactions handled elsewhere
  else if (grepl("landgebruik", s)) NA_character_     # factor levels
  else s
}

cond_tab$base <- vapply(cond_tab$term, base_var, character(1))

# compute IQRs from the modelling data (use the same frame used in m_richness_full)
df_mod <- insight::get_data(top_models[[1]])  # or the data you passed to glmmTMB

iqr_map <- df_mod %>%
  summarize(
    across(any_of(unique(na.omit(cond_tab$base))),
           ~ IQR(.x, na.rm = TRUE), .names = "iqr_{col}")
  ) %>% tidyr::pivot_longer(everything(),
       names_to = "base", values_to = "IQR") %>%
  mutate(base = sub("^iqr_", "", base))

cond_cont <- cond_tab %>%
  left_join(iqr_map, by = "base") %>%
  filter(!is.na(base), !grepl("^poly\\(", term), !grepl(":", term)) %>%
  mutate(
    RR_IQR     = exp(estimate * IQR),
    RR_IQR_lwr = exp(lwr * IQR),
    RR_IQR_upr = exp(upr * IQR),
    pct_change_IQR = (RR_IQR - 1) * 100
  )

```

Categorical (land use) → ratios vs reference

```{r}
cond_cat <- cond_tab %>%
  filter(grepl("^landgebruik", term)) %>%
  mutate(
    RR_unit  = exp(estimate),
    RR_lwr   = exp(lwr),
    RR_upr   = exp(upr),
    pct_vs_ref = (RR_unit - 1) * 100
  )

```


Plot (response scale, model-averaged):

Continuous (IQR % change)

```{r}
ggplot(cond_cont, aes(x = reorder(term, pct_change_IQR), y = pct_change_IQR)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_pointrange(aes(ymin = (RR_IQR_lwr - 1)*100, ymax = (RR_IQR_upr - 1)*100)) +
  coord_flip() + theme_minimal() +
  labs(x = "Predictor", y = "% change in richness per IQR (model-averaged)",
       title = "Model-averaged effects (ΔAICc set)")
```

Categorical (vs reference)

```{r}
ggplot(cond_cat, aes(x = term, y = pct_vs_ref)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_pointrange(aes(ymin = (RR_lwr - 1)*100, ymax = (RR_upr - 1)*100)) +
  coord_flip() + theme_minimal() +
  labs(x = "Land use level (vs ref)", y = "% difference in richness",
       title = "Model-averaged contrasts (ΔAICc set)")
```


##### Model met de laagste AICc:

```{r}
lowest_aic <- get.models(dd, 1)[[1]]
summary(lowest_aic)
```

```{r m-richness-anova-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
car::Anova(lowest_aic) |> print(digits = 2)
```

```{r m-richness-marginal-means-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
# Estimated marginal means per land use *within each depth layer*
emmeans_landuse_by_depth_incl_fyschem <- marginaleffects::avg_predictions(
  lowest_aic,
  variables = "landgebruik",
  newdata = data_subset,
  by = c("diepte", "landgebruik"),
  type = "response"
)


# Save to CSV with automatic naming
write_csv(
  emmeans_landuse_by_depth_incl_fyschem,
  file = file.path(output_dir, glue("emmeans_landuse_by_depth_incl_fyschem_{group}_{primerset}_{unit}.csv"))
)

# print
emmeans_landuse_by_depth_incl_fyschem
```

```{r m-richness-marginaleffects-contrasts-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
# Pairwise contrasts between land use types *within each depth layer*
contrasts_landuse_by_depth <- marginaleffects::comparisons(
  model = lowest_aic,
  variables = "landgebruik",
  newdata = data_subset,
  type = "response",
  by = "diepte")

# Save to CSV with automatic naming
write_csv(
  contrasts_landuse_by_depth,
  file = file.path(output_dir, glue("margeffects_contrasts_landuse_by_depth_incl_fyschem_{group}_{primerset}_{unit}.csv"))
)

# print
contrasts_landuse_by_depth
```

```{r m-richness-contrasts-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
# Estimate marginal means per land use per depth
emm_landuse_by_depth <- emmeans(
  lowest_aic,
  specs = "landgebruik",
  by = "diepte",
  data = data_subset
)

# Compute pairwise contrasts
contrasts_landuse_by_depth_incl_fyschem <- contrast(emm_landuse_by_depth, method = "pairwise", adjust = "none", infer = c(TRUE, TRUE)) %>%
  as.data.frame()

# Save to CSV
write_csv(
  contrasts_landuse_by_depth_incl_fyschem,
  file = file.path(output_dir, glue("contrasts_landuse_by_depth_incl_fyschem_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_landuse_by_depth_incl_fyschem)

```

```{r m-richness-contrasts-incl-fyschem-adj-{{group}}-{{primerset}}-{{unit}}}
# Add bh-corrected p-values per depth group
contrasts_landuse_by_depth_incl_fyschem_adj <- contrasts_landuse_by_depth_incl_fyschem %>%
  as_tibble() %>%
  group_by(diepte) %>%
  mutate(p_bh = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  select(
    diepte,
    contrast,
    estimate,
    SE,
    z.ratio,
    p.value,
    p_bh,
    asymp.LCL,
    asymp.UCL
  ) %>%
  arrange(diepte, p.value)

# Save adjusted results to CSV
write_csv(
  contrasts_landuse_by_depth_incl_fyschem_adj,
  file = file.path(output_dir, glue("contrasts_landuse_by_depth_incl_fyschem_adj_{group}_{primerset}_{unit}.csv"))
)

# Print adjusted summary
print(contrasts_landuse_by_depth_incl_fyschem_adj)

```


```{r error=TRUE}
p <- performance::check_predictions(lowest_aic)
plot(p) + facet_zoom(xlim = c(0, 10))
p <- performance::check_overdispersion(lowest_aic)
p
p %>% plot()
p <- performance::check_residuals(lowest_aic)
p
p %>% plot()
performance::check_collinearity(lowest_aic)
performance::check_zeroinflation(lowest_aic)
```


```{r}
at <- car::Anova(lowest_aic)
efvars <- rownames(at)[order(-at$Chisq)]
interactingvars <- efvars[grepl(":", efvars)]
interactingvars <- stringr::str_split(interactingvars, ":") %>%
  unlist() %>%
  unique()
efvars <- efvars[!efvars %in% interactingvars]
efvars <- stringr::str_replace_all(
  efvars, "^(\\w+\\()?(.+?)(\\))?$", "\\2"
)
efvars <- ifelse(
  efvars == "diepte:landgebruik", "landgebruik:diepte", efvars
)
efvars <- ifelse(
  efvars == "total_count + 1", "total_count", efvars
)
efvars <- ifelse(
  efvars == "ph_kcl, 2", "ph_kcl", efvars
)

p <- vector(mode = "list", length = length(efvars))
p <- set_names(p, efvars)
for (i in efvars) {
  if (grepl(":", i)) {
    conds <- stringr::str_split_1(i, ":")
    nd <- insight::get_datagrid(
      lowest_aic,
      by = conds,
      length = 50
    )
    pl <- marginaleffects::plot_predictions(
      model = lowest_aic,
      by = conds,
      newdata = nd,
      vcov = TRUE,
      type = "response",
      points = 0.1,
      re.form = NA
    )
  } else {
    nd <- insight::get_datagrid(
      lowest_aic,
      by = i,
      length = 50
    )
    pl <- marginaleffects::plot_predictions(
      lowest_aic,
      by = i,
      newdata = nd,
      vcov = TRUE,
      type = "response",
      points = 0.1,
      re.form = NA
    )
  }
  p[[i]] <- pl
}

p
```
