### `r stringr::str_to_sentence("{{group}} - {{primerset}} - {{unit}}")`

```{r}
data_subset <- combined %>%
  filter(
    group == "{{group}}",
    primerset == "{{primerset}}",
    unit == "{{unit}}"
  )
```


```{r m-richness-{{group}}-{{primerset}}-{{unit}}}
form_richness <- formula(
  observed ~
    log(total_count)
  + landgebruik
  + diepte
  + landgebruik:diepte
  + (1 | cmon_plot_id)
  )

cat("Fitting Poisson model")

m_richness <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = poisson())

test <- check_overdispersion(m_richness)
test <- test$p_value < 0.05

if (test) {
  cat("Overdispersion detected: fitting Negative binomial model instead.")
  m_richness <- glmmTMB(
    formula = form_richness,
    data = data_subset,
    family = nbinom2())
}
```


```{r m-richness-checks-{{group}}-{{primerset}}-{{unit}}}
performance::check_model(m_richness)
```


```{r m-richness-summary-{{group}}-{{primerset}}-{{unit}}}
summary(m_richness) |> print(digits = 2)
```

```{r m-richness-anova-{{group}}-{{primerset}}-{{unit}}}
car::Anova(m_richness) |> print(digits = 2)
```

```{r m-richness-preds-total-count-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("total_count"),
  re.form = NA,
  vcov = TRUE,
  type = "response") +
  labs(y = "Voorspeld aantal taxa") +
  scale_x_log10()
```

```{r m-richness-preds-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("landgebruik", "diepte"),
  re.form = NA,
  vcov = TRUE,
  type = "response") +
  labs(y = "Voorspeld aantal taxa")
```


### `r stringr::str_to_sentence("{{group}} - {{primerset}} - {{unit}} including SWCvol")`

Including soil water content (SWCvol) results in a slightly better fit overall (based on the AIC, BIC, log-likelihood, and deviance values)? However, SWCvol and its interaction with depth do not significantly improve the model (p > 0.05)?

Removing the interaction between SWCvol and depth results in a model with slightly better AIC and BIC values than the model that includes the interaction, and the p-value for SWCvol has also improved slightly, but it is still not significant.


```{r m-richness-incl-SWCvol-{{group}}-{{primerset}}-{{unit}}}
form_richness <- formula(
  observed ~
    log(total_count)
  + Landgebruik_MBAG
  + Diepte
  + SWCvol
  + Landgebruik_MBAG:Diepte
  + SWCvol:Diepte
  + (1 | Cmon_PlotID)
  )

cat("Fitting Poisson model")

m_richness <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = poisson())

test <- check_overdispersion(m_richness)
test <- test$p_value < 0.05

if (test) {
  cat("Overdispersion detected: fitting Negative binomial model instead.")
  m_richness <- glmmTMB(
    formula = form_richness,
    data = data_subset,
    family = nbinom2())
}
```


```{r m-richness-incl-SWCvol-checks-{{group}}-{{primerset}}-{{unit}}}
performance::check_model(m_richness)
```


```{r m-richness-incl-SWCvol-summary-{{group}}-{{primerset}}-{{unit}}}
summary(m_richness) |> print(digits = 2)
```

```{r m-richness-incl-SWCvol-anova-{{group}}-{{primerset}}-{{unit}}}
car::Anova(m_richness) |> print(digits = 2)
```

```{r m-richness-incl-SWCvol-preds-total-count-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("total_count"),
  re.form = NA,
  vcov = TRUE,
  type = "response") +
  labs(y = "Voorspeld aantal taxa") +
  scale_x_log10()
```

```{r m-richness-incl-SWCvol-preds-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("landgebruik", "diepte"),
  re.form = NA,
  vcov = TRUE,
  type = "response") +
  labs(y = "Voorspeld aantal taxa")
```




