### `r stringr::str_to_sentence("{{group}} - {{primerset}} - {{unit}}")`

```{r}
group <- "{{group}}"
primerset <- "{{primerset}}"
unit <- "{{unit}}"
season_run <- "{{season_run}}"

```

Seasonal filtering inside the child

```{r}
# Seasonal filtering inside the child
months_map <- list(
  allseasons     = NULL,
  winterhalfjaar = c(11, 12, 1, 2, 3, 4),
  winter         = c(12, 1, 2, 3, 4)
)

data_subset <- combined %>%
  {
    sel <- months_map[[season_run]]
    if (!is.null(sel) && "sampling_month" %in% names(.)) {
      dplyr::filter(., sampling_month %in% sel)
    } else {
      .
    }
  } %>%
  dplyr::filter(
    group     == "{{group}}",
    primerset == "{{primerset}}",
    unit      == "{{unit}}"
  ) %>%
  droplevels()
```

```{r}
cat("season_run =", season_run,
    "| rows =", nrow(data_subset),
    "| months =",
    if ("sampling_month" %in% names(data_subset))
      paste(sort(unique(data_subset$sampling_month)), collapse = ",")
    else "NA",
    "\n")

```

```{r}
data_subset %>%
  group_by(landgebruik, diepte) %>%
  summarize(n_shannon_not_missing = sum(!is.na(shannon))) %>%
  knitr::kable()
```


```{r}
# remove NAs
data_subset <- data_subset %>%
  filter(!is.na(shannon))

diepte_ok <- length(unique(data_subset$diepte)) > 1
```


```{r m-shannon-{{group}}-{{primerset}}-{{unit}}}
if (diepte_ok) {
  form_shannon <- formula(
  shannon ~
    log(total_count + 1)
    + landgebruik
    + diepte
    + landgebruik:diepte
    + (1 | cmon_plot_id)
  )
} else {
  form_shannon <- formula(
  shannon ~
    log(total_count + 1)
    + landgebruik
  + (1 | cmon_plot_id)
  )
}

zeroes <- min(data_subset$shannon, na.rm = TRUE) == 0

if (!zeroes) {
  cat("Fitting Gamma model")

  m_shannon <- glmmTMB(
    formula = form_shannon,
    data = data_subset,
    family = Gamma(link = "log")
  )
} else {
  cat(
    "Zeroes detected (only one taxum observed):
    Fitting zero-inflated Gamma model"
  )

  m_shannon <- glmmTMB(
    formula = form_shannon,
    ziformula = ~1,
    data = data_subset,
    family = ziGamma(link = "log")
  )
}
```


```{r m-shannon-checks-{{group}}-{{primerset}}-{{unit}}}
performance::check_model(m_shannon)
```


```{r m-shannon-summary-{{group}}-{{primerset}}-{{unit}}}
summary(m_shannon) |> print(digits = 2)
```

Shannon: capture model family

```{r}
# ---- Shannon: capture model family (NEW) ----
get_family_name <- function(model) {
  tryCatch({
    fam <- stats::family(model)
    if (is.list(fam) && !is.null(fam$family)) as.character(fam$family)
    else if (inherits(fam, "family") && !is.null(fam$family)) fam$family
    else if (is.character(fam)) fam
    else NA_character_
  }, error = function(e) NA_character_)
}

model_family_shannon <- get_family_name(m_shannon)

append_model_family_shannon(
  group        = group,
  primerset    = primerset,
  season_run   = season_run,
  model_family = model_family_shannon
)

```


```{r m-shannon-anova-{{group}}-{{primerset}}-{{unit}}}
car::Anova(m_shannon) |> print(digits = 2)
```

estimate contrasts between land-use categories per depth

```{r m-shannon-emmeans-contrasts-per-depth-{{group}}-{{primerset}}-{{unit}}}
# Estimate marginal means per land use per depth
emm_shannon_landuse_by_depth <- emmeans(
  m_shannon,
  specs = "landgebruik",
  by = "diepte",
  data = data_subset,
  type = "response"
)

# Pairwise contrasts (auto adjustment by emmeans; keep CIs & p-values)
contrasts_shannon_landuse_response <- contrast(
  emm_shannon_landuse_by_depth,
  method = "pairwise",
  infer = c(TRUE, TRUE)
) %>%
  as.data.frame()

# Save to CSV
write_csv(
  contrasts_shannon_landuse_response,
  file = file.path(output_dir, glue("contrasts_landuse_response_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_shannon_landuse_response)

```

accumulator contrasts between land-use categories per depth

```{r}
# ---- Append to parent accumulator (NEW) ----
contrasts_tbl_sh <- contrasts_shannon_landuse_response %>%
  tibble::as_tibble() %>%
  # ensure diepte column exists (it should, since by="diepte")
  { if (!"diepte" %in% names(.)) dplyr::mutate(., diepte = NA_character_) else . } %>%
  # split "Akker / Natuurgrasland" into two columns
  tidyr::separate(.data$contrast,
                  into = c("landgebruik_1","landgebruik_2"),
                  sep  = " / ",
                  remove = TRUE, fill = "right") %>%
  # harmonize columns from emmeans outputs
  dplyr::mutate(
    estimate_diff = dplyr::coalesce(.data$estimate, .data$contrast),  # differences on response scale
    SE            = .data$SE,
    lcl           = dplyr::coalesce(.data$asymp.LCL, .data$lower.CL),
    ucl           = dplyr::coalesce(.data$asymp.UCL, .data$upper.CL),
    stat          = dplyr::coalesce(.data$z.ratio, .data$t.ratio),
    stat_name     = dplyr::case_when(!is.na(.data$z.ratio) ~ "z",
                                     !is.na(.data$t.ratio) ~ "t",
                                     TRUE ~ NA_character_)
  ) %>%
  # tag with identifiers
  dplyr::mutate(
    group      = group,
    primerset  = primerset,
    season_run = season_run
  ) %>%
  dplyr::select(group, primerset, season_run, diepte,
                landgebruik_1, landgebruik_2,
                estimate_diff, SE, lcl, ucl, stat_name, stat, p.value)

append_summary_contrasts_shannon(contrasts_tbl_sh)


```

marginal effects average predictions (double check marginal means but different method)

```{r m-shannon-preds-values-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
predicted_values <- marginaleffects::avg_predictions(
  m_shannon,
  variables = c("landgebruik", "diepte"),
  newdata = data_subset,
  type = "response"
)

# save to CSV
write_csv(
  predicted_values,
  file = file.path(output_dir, glue("predicted_values_shannon_{group}_{primerset}_{unit}.csv"))
)

# View
print(predicted_values)
```

Plot predictions total count

```{r m-shannon-preds-total-count-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_shannon,
  condition = c("total_count"),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Shannon index") +
  scale_x_log10()
```

plot predictions design variables

```{r m-shannon-preds-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_shannon,
  condition = c("landgebruik", "diepte"[diepte_ok]),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Shannon index")
```


