
### `r stringr::str_to_sentence("{{group}} - {{primerset}} - {{unit}}")`

```{r}
group <- "{{group}}"
primerset <- "{{primerset}}"
unit <- "{{unit}}"
```

```{r}
data_subset <- combined %>%
  filter(
    group == "{{group}}",
    primerset == "{{primerset}}",
    unit == "{{unit}}"
  )
```


```{r}
data_subset %>%
  group_by(landgebruik, diepte) %>%
  summarize(n_observed_not_missing = sum(!is.na(observed))) %>%
  knitr::kable()
```

```{r}
# remove NAs
data_subset <- data_subset %>%
  filter(!is.na(observed))

diepte_ok <- length(unique(data_subset$diepte)) > 1
```


#### Basismodel

```{r m-richness-{{group}}-{{primerset}}-{{unit}}}
if (diepte_ok) {
  form_richness <- formula(
  observed ~
    log(total_count + 1)
    + landgebruik
    + diepte
    + landgebruik:diepte
    + (1 | cmon_plot_id)
  )
  ziform <- formula(~ 1 + (1 | cmon_plot_id))
} else {
  form_richness <- formula(
  observed ~
    log(total_count + 1)
    + landgebruik
  )
  ziform <- formula(~ 1)
}

cat("Fitting Poisson model\n")
m_pois <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = poisson()
)
cat("Fitting hurdle Poisson model\n")
m_hpois <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_poisson()
)
cat("Fitting negative binomial model\n")
m_nbinom <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = nbinom2()
)
cat("Fitting hurdle negative binomial model\n")
m_hnbinom <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_nbinom2()
)
cat("Fitting generalized Poisson model\n")
m_genpois <- glmmTMB(
  formula = form_richness,
  data = data_subset,
  family = genpois()
)
cat("Fitting hurdle generalized Poisson model\n")
m_hgenpois <- glmmTMB(
  formula = form_richness,
  ziformula = ziform,
  data = data_subset,
  family = truncated_genpois()
)

mlist <- list(
  pois = m_pois,
  hpois = m_hpois,
  nbinom = m_nbinom,
  hnbinom = m_hnbinom,
  genpois = m_genpois,
  hgenpois = m_hgenpois
)
```

Het model met de laagste AIC waarde (hoogste AIC gewicht: `AIC_wt`) heeft de beste fit volgens AIC criterium.
Dit betekent niet noodzakelijk dat dit model geen enkel probleem meer heeft met betrekking tot `zero-inflation` (`deflation`) of over- (onder-) dispersie.

Modellen waarvan de AIC niet berekend kan worden, worden verwijderd.
Deze modellen zijn wellicht niet geconvergeerd.

```{r}
converged <- !is.na(map_dbl(mlist, AIC))
cp <- compare_performance(mlist[converged], metrics = c("AIC"), rank = FALSE) |>
  bind_cols(map_dfr(mlist[converged], inflation_dispersion))
kable(cp, digits = 2)
```

```{r}
m_richness <- mlist[[cp$Name[cp$AIC_wt == max(cp$AIC_wt)]]]
```


Het geselecteerde model is:

```{r}
insight::get_call(m_richness)
```

```{r m-richness-checks-{{group}}-{{primerset}}-{{unit}}, error=TRUE}
p <- performance::check_predictions(m_richness)
plot(p) + facet_zoom(xlim = c(0, 10))
p <- performance::check_overdispersion(m_richness)
p
p %>% plot()
p <- performance::check_residuals(m_richness)
p
p %>% plot()
performance::check_collinearity(m_richness)
performance::check_zeroinflation(m_richness)
```


```{r m-richness-summary-{{group}}-{{primerset}}-{{unit}}}
summary(m_richness) |> print(digits = 2)
```

```{r m-richness-anova-{{group}}-{{primerset}}-{{unit}}}
car::Anova(m_richness) |> print(digits = 2)
```

```{r m-richness-marginal-means-per-depth-{{group}}-{{primerset}}-{{unit}}}
# Estimated marginal means per land use (and per depth if applicable)

emm_landuse <- emmeans(
  m_richness,
  specs = "landgebruik",
  by = if (diepte_ok) "diepte" else NULL,
  data = data_subset,
  type = "response"
)

# Save to CSV with automatic naming
write_csv(
  as.data.frame(emm_landuse),
  file = file.path(output_dir, glue("emmeans_landuse_{group}_{primerset}_{unit}.csv"))
)

# print
emm_landuse


```


```{r m-richness-contrasts-per-depth-response-{{group}}-{{primerset}}-{{unit}}}
# Estimated marginal means per land use (and per depth if applicable), on response scale
emm_landuse <- emmeans(
  m_richness,
  specs = "landgebruik",
  by = if (diepte_ok) "diepte" else NULL,
  data = data_subset,
  type = "response"
)

# Pairwise contrasts (auto adjustment by emmeans; keep CIs & p-values)
contrasts_landuse_response <- contrast(
  emm_landuse,
  method = "pairwise",
  infer = c(TRUE, TRUE)
) %>%
  as.data.frame()

# Save to CSV
write_csv(
  contrasts_landuse_response,
  file = file.path(output_dir, glue("contrasts_landuse_response_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_landuse_response)


```


```{r m-richness-contrasts-per-depth-response-bh-{{group}}-{{primerset}}-{{unit}}}
# Summary table
contrasts_landuse_table <- contrasts_landuse_response %>%
  as_tibble() %>%
  dplyr::select(
    dplyr::any_of(c(
      "diepte",        # only present if diepte_ok == TRUE
      "contrast",
      "estimate",      # default column name
      "ratio",         # appears instead of 'estimate' when type = "response"
      "SE",
      "z.ratio",       # or t.ratio, one of them will always exist with infer=TRUE
      "p.value",
      "asymp.LCL",
      "asymp.UCL"
    ))
  )

# Sort
if (diepte_ok) {
  contrasts_landuse_table <- contrasts_landuse_table %>% arrange(diepte, p.value)
} else {
  contrasts_landuse_table <- contrasts_landuse_table %>% arrange(p.value)
}

# Save
write_csv(
  contrasts_landuse_table,
  file = file.path(output_dir, glue("contrasts_landuse_table_response_{group}_{primerset}_{unit}.csv"))
)

# Print
contrasts_landuse_table

```

```{r m-richness-preds-total-count-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("total_count"),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Voorspeld aantal taxa")
```

```{r m-richness-preds-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_richness,
  condition = c("landgebruik", "diepte"[diepte_ok]),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Voorspeld aantal taxa")
```


#### Met fysicochemische data

Stapsgewijze modelselectie:

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "c_density",
  "swc_vol",
  "ph_kcl",
  "textuur_som_klei_zand"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset[
  complete.cases(data_subset[, variables_no_collinearity]),
]
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```


We voegen alle niet collineaire fysicochemische variabelen toe aan het basismodel.
Voor pH gebruiken we een kwadratische effect.
Dit wordt het `full model`.

```{r extend-model-additional-predictors-{{group}}-{{primerset}}-{{unit}}}
form_richness_full <- update(
  form_richness,
  . ~ .
    + swc_vol
    + c_density
    + poly(ph_kcl, 2)
    + textuur_som_klei_zand
)

m_richness_full <- update(
  m_richness,
  formula = form_richness_full,
  data = data_subset,
  na.action = na.fail
)
```

We vergelijken de parameterschattingen van het basis en volledig model.

```{r coefficients-se-{{group}}-{{primerset}}-{{unit}}}
base_model_coefs <- broom.mixed::tidy(m_richness)
full_model_coefs <- broom.mixed::tidy(m_richness_full)

coefs_df <- rbind(
  data.frame(Model = "Base", base_model_coefs),
  data.frame(Model = "Full", full_model_coefs)
)
```

```{r plot-coefficient-estimates-{{group}}-{{primerset}}-{{unit}}}
ggplot(
  coefs_df,
  aes(x = term, y = estimate, fill = Model)
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(
    aes(
      ymin = estimate - std.error,
      ymax = estimate + std.error
    ),
    width = 0.2, position = position_dodge(0.9)
  ) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Coefficient Estimates and Standard Errors",
    x = "Predictor",
    y = "Estimate",
    fill = "Model"
  )
```

```{r save-coefficient-estimates-{{group}}-{{primerset}}-{{unit}}}
readr::write_csv(
  coefs_df,
  file = file.path(
    output_dir,
    glue::glue("coefs_base_vs_full_{group}_{primerset}_{unit}.csv")
  )
)
```


Modelselectie met de functie `dredge`.
We beperken het aantal te verkennen parametercombinaties door te forceren dat `total count` `landgebruik` en `diepte` altijd geschat moeten worden ('by design') en maximaal 7 termen in het model zitten.

```{r model-selection-and-comparison-{{group}}-{{primerset}}-{{unit}}, warning=FALSE}
if (diepte_ok) {
  model_set <- dredge(
    m_richness_full,
    m.lim = c(NA, 7),
    fixed = ~
      cond(log(total_count + 1))
    + cond(landgebruik)
    + cond(diepte)
    + (1 | cmon_plot_id)
  )
} else {
  model_set <- dredge(
    m_richness_full,
    m.lim = c(NA, 7),
    fixed = ~
      cond(log(total_count + 1))
    + cond(landgebruik)
    + (1 | cmon_plot_id)
  )
}
```

Per variabele berekenen we de som van modelgewichten over alle modellen en tellen we in hoeveel modellen de variabele voorkwam:

```{r importance-per-predictor-{{group}}-{{primerset}}-{{unit}}}
importance_df <- sw(model_set)
print(importance_df)
```

Visualisatie van modellen die minder dan 4 AICc verschillen van het model met laagste AICc (rij 1):

```{r}
dd <- subset(model_set, delta < 4)

par(mar = c(3, 5, 6, 4))
plot(dd, labAsExpr = TRUE)
```

```{r save-model-selection-{{group}}-{{primerset}}-{{unit}}}

# -- Full selection table
model_set_tbl <- model_set %>%
  as.data.frame() %>%
  rownames_to_column("model_id") %>%
  arrange(AICc)

write_csv(
  model_set_tbl,
  file.path(output_dir, glue("full_model_set_{group}_{primerset}_{unit}.csv"))
)

# -- ΔAICc < 4 subset
dd_obj <- subset(model_set, delta < 4)
dd_tbl <- dd_obj %>%
  as.data.frame() %>%
  rownames_to_column("model_id") %>%
  arrange(AICc)

write_csv(
  dd_tbl,
  file.path(output_dir, glue("model_set_delta_lt4_{group}_{primerset}_{unit}.csv"))
)

# -- 95% cumulative weight subset
cumw_tbl <- model_set_tbl %>%
  mutate(cum_weight = cumsum(weight)) %>%
  filter(cum_weight <= 0.95)

write_csv(
  cumw_tbl,
  file.path(output_dir, glue("model_set_cumw95_{group}_{primerset}_{unit}.csv"))
)

# Variable importance (sum of weights by term)
imp <- MuMIn::importance(model_set)
imp_tbl <- tibble(term = names(imp), importance = as.numeric(imp)) %>%
  arrange(desc(importance))

write_csv(
  imp_tbl,
  file.path(output_dir, glue("model_importance_{group}_{primerset}_{unit}.csv"))
)

# -- (Reproducibility) Save the objects too
saveRDS(model_set, file.path(output_dir, glue("model_set_{group}_{primerset}_{unit}.rds")))
saveRDS(dd_obj,    file.path(output_dir, glue("model_set_delta_lt4_{group}_{primerset}_{unit}.rds")))


```


Dit is de samenvatting voor het model met de laagste AICc:

```{r}
lowest_aic <- get.models(dd, 1)[[1]]
summary(lowest_aic)
```

```{r m-richness-anova-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
car::Anova(lowest_aic) |> print(digits = 2)
```

```{r m-richness-marginal-means-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
# Estimated marginal means per land use (and per depth if applicable), on response scale
emm_landuse_incl_fyschem <- emmeans(
  lowest_aic,
  specs = "landgebruik",
  by    = if (diepte_ok) "diepte" else NULL,
  data  = data_subset,
  type  = "response"
)

# Keep original filename
write_csv(
  as.data.frame(emm_landuse_incl_fyschem),
  file = file.path(output_dir, glue("emmeans_landuse_incl_fyschem_{group}_{primerset}_{unit}.csv"))
)

# Print
emm_landuse_incl_fyschem

```

```{r m-richness-contrasts-per-depth-response-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
# Pairwise contrasts on response scale (keep CIs & p-values), via emmeans
contrasts_landuse_response_incl_fyschem <- contrast(
  emm_landuse_incl_fyschem,
  method = "pairwise",
  infer  = c(TRUE, TRUE)
) %>%
  as.data.frame()

# Write
write_csv(
  contrasts_landuse_response_incl_fyschem,
  file = file.path(output_dir, glue("contrasts_landuse_incl_fyschem_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_landuse_response_incl_fyschem)

```


```{r m-richness-contrasts-incl-fyschem-adj-{{group}}-{{primerset}}-{{unit}}}
# Summary table, aligned with base-model structure
contrasts_landuse_by_depth_incl_fyschem_adj <- contrasts_landuse_response_incl_fyschem %>%
  as_tibble() %>%
  dplyr::select(
    dplyr::any_of(c(
      "diepte",        # only present if diepte_ok == TRUE
      "contrast",
      "estimate",      # link-scale (absent when type = "response")
      "ratio",         # present when type = "response" (IRR)
      "SE",
      "z.ratio",       # or t.ratio
      "p.value",
      "asymp.LCL",
      "asymp.UCL"
    ))
  ) %>%
  { if (diepte_ok) arrange(., diepte, p.value) else arrange(., p.value) }

# Keep original adjusted filename (content is now the summary table without BH)
write_csv(
  contrasts_landuse_by_depth_incl_fyschem_adj,
  file = file.path(output_dir, glue("contrasts_landuse_incl_fyschem_adj_{group}_{primerset}_{unit}.csv"))
)

# Print
contrasts_landuse_by_depth_incl_fyschem_adj

```


```{r error=TRUE}
p <- performance::check_predictions(lowest_aic)
plot(p) + facet_zoom(xlim = c(0, 10))
p <- performance::check_overdispersion(lowest_aic)
p
p %>% plot()
p <- performance::check_residuals(lowest_aic)
p
p %>% plot()
performance::check_collinearity(lowest_aic)
performance::check_zeroinflation(lowest_aic)
```


```{r}
at <- car::Anova(lowest_aic)
efvars <- rownames(at)[order(-at$Chisq)]
interactingvars <- efvars[grepl(":", efvars)]
interactingvars <- stringr::str_split(interactingvars, ":") %>%
  unlist() %>%
  unique()
efvars <- efvars[!efvars %in% interactingvars]
efvars <- stringr::str_replace_all(
  efvars, "^(\\w+\\()?(.+?)(\\))?$", "\\2"
)
efvars <- ifelse(
  efvars == "diepte:landgebruik", "landgebruik:diepte", efvars
)
efvars <- ifelse(
  efvars == "total_count + 1", "total_count", efvars
)
efvars <- ifelse(
  efvars == "ph_kcl, 2", "ph_kcl", efvars
)

p <- vector(mode = "list", length = length(efvars))
p <- set_names(p, efvars)
for (i in efvars) {
  if (grepl(":", i)) {
    conds <- stringr::str_split_1(i, ":")
    nd <- insight::get_datagrid(
      lowest_aic,
      by = conds,
      length = 50
    )
    pl <- marginaleffects::plot_predictions(
      model = lowest_aic,
      by = conds,
      newdata = nd,
      vcov = TRUE,
      type = "response",
      points = 0.1,
      re.form = NA
    )
  } else {
    nd <- insight::get_datagrid(
      lowest_aic,
      by = i,
      length = 50
    )
    pl <- marginaleffects::plot_predictions(
      lowest_aic,
      by = i,
      newdata = nd,
      vcov = TRUE,
      type = "response",
      points = 0.1,
      re.form = NA
    )
  }
  p[[i]] <- pl
}

p
```

global c_density effect (IRR per IQR), with 95% CI. This gives a global (averaged) slope across the different land-use categories

```{r}
# Global IRR per IQR for c_density from the additive model
iqr_cd <- IQR(data_subset$c_density, na.rm = TRUE)
tr_add <- summary(emtrends(lowest_aic, ~ 1, var = "c_density", mode = "latent", infer = TRUE))
irr_add <- transform(as.data.frame(tr_add),
                     increment = paste0("per IQR (", round(iqr_cd,2), ")"),
                     IRR = exp(c_density.trend * iqr_cd),
                     LCL = exp((c_density.trend - 1.96*SE) * iqr_cd),
                     UCL = exp((c_density.trend + 1.96*SE) * iqr_cd))
irr_add[, c("increment","IRR","LCL","UCL","p.value")]
```


#### Correlation carbon density per land-use type

We voegen carbon density toe aan het basismodel.

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "c_density"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset[
  complete.cases(data_subset[, variables_no_collinearity]),
]
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```


```{r extend-model-carbon-density-{{group}}-{{primerset}}-{{unit}}}
form_richness_full <- update(
  form_richness,
  . ~ .
    + c_density)

m_richness_full <- update(
  m_richness,
  formula = form_richness_full,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
summary(m_richness_full)
```

We voegen carbon density en de interactie met landgebruik toe aan het basismodel (to see land-use–specific slopes).

```{r}
form_richness_full_int <- update(
  form_richness,
  . ~ .
    + landgebruik * c_density)

m_richness_full_int <- update(
  m_richness,
  formula = form_richness_full_int,
  data = data_subset,
  na.action = na.fail
)
```

```{r}
m_richness_full_int

summary(m_richness_full_int)

```

```{r}
anova(m_richness_full, m_richness_full_int)  # LRT for interaction

```


Report per–land-use slopes of c_density (IRR per IQR) without subsetting

```{r}
iqr_cd <- IQR(data_subset$c_density, na.rm = TRUE)
emtrends(m_richness_full_int, ~ landgebruik, var = "c_density",
         delta = iqr_cd, type = "response", infer = TRUE)

```


plot

Diagnostics for the interaction model

```{r}

mod <- m_richness_full_int

# Predictions vs observed (zoom to data-driven x-range if ggforce available)
p1 <- performance::check_predictions(mod)
xr <- quantile(data_subset$c_density, c(.01, .99), na.rm = TRUE)

if (requireNamespace("ggforce", quietly = TRUE)) {
  library(ggforce)
  print(plot(p1) + ggforce::facet_zoom(xlim = xr))
} else {
  print(plot(p1))
}

# Overdispersion, residuals, collinearity, zero-inflation
od <- performance::check_overdispersion(mod); print(od); plot(od)
res <- performance::check_residuals(mod);    print(res); plot(res)
performance::check_collinearity(mod)
performance::check_zeroinflation(mod)

```

Model-based predictions: richness vs c_density by landgebruik

```{r}
# Build grid over c_density within each land-use (keep your “pinning” lines)
by_vars <- c("landgebruik", "c_density")
nd <- insight::get_datagrid(mod, by = by_vars, length = 50)

if ("total_count" %in% names(data_subset)) {
  nd$total_count <- stats::median(data_subset$total_count, na.rm = TRUE)
}
ref_depth <- if ("diepte" %in% names(data_subset) &&
                 any(grepl("0-10", levels(data_subset$diepte)))) "0-10" else levels(data_subset$diepte)[1]
nd$diepte <- ref_depth

# Get predictions (response scale), marginal over REs
pr <- marginaleffects::predictions(
  model   = mod,
  newdata = nd,
  type    = "response",
  vcov    = TRUE,
  re.form = NA
)

# Plot: x = c_density, facet by land-use
library(ggplot2)
ggplot(pr, aes(x = c_density, y = estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(x = "Carbon density", y = "Predicted richness",
       title = "Predicted richness vs carbon density by land-use",
       subtitle = paste0("Predictions at total_count = ", round(stats::median(data_subset$total_count, na.rm = TRUE)),
                         ", depth = ", ref_depth)) +
  theme_bw()


```

```{r}
resp_var <- all.vars(formula(mod))[1]
last_plot() +
  geom_point(data = data_subset,
             aes(x = c_density, y = .data[[resp_var]]),
             inherit.aes = FALSE, alpha = 0.12, size = 0.9)

```

alternative plot option

```{r}
# Universal predictions + facet by land-use
library(insight)
library(marginaleffects)
library(ggplot2)

# Build grid over c_density within each land-use
by_vars <- c("landgebruik", "c_density")
nd <- insight::get_datagrid(mod, by = by_vars, length = 50)

# Pin nuisance covariates for standardized curves (optional)
if ("total_count" %in% names(nd)) {
  nd$total_count <- stats::median(data_subset$total_count, na.rm = TRUE)
}
if ("diepte" %in% names(nd)) {
  ref_depth <- if (any(grepl("0-10", levels(droplevels(data_subset$diepte)))))
    "0-10" else levels(droplevels(data_subset$diepte))[1]
  nd$diepte <- ref_depth
}

# Predict on the response scale; exclude random effects for marginal curves
pr <- predictions(
  model   = mod,
  newdata = nd,
  type    = "response",
  re.form = NA   # for GLMMs; harmless for GLMs
)

# Plot: x = c_density, facet by land-use
ggplot(pr, aes(x = c_density, y = estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(x = "Carbon density", y = "Predicted richness") +
  theme_bw()


```

```{r}
library(ggeffects)
library(insight)

gg_type <- function(mod) {
  mi <- insight::model_info(mod)
  # Zero-inflated models (glmmTMB, etc.)
  if (isTRUE(mi$is_zero_inflated)) {
    return("zero_inflated")  # conditional mean (mu). Use "zi_prob" to plot pi.
  }
  # Count models
  if (isTRUE(mi$is_count)) {
    return("count")
  }
  # Survival models
  if (isTRUE(mi$is_survival)) {
    return("survival")  # or "cumulative_hazard" depending on what you want
  }
  # Default fallback (works for Gaussian, binomial, etc.)
  "fixed"
}

tp <- gg_type(mod)

pred <- ggpredict(
  mod,
  terms = c("c_density [all]", "landgebruik"),
  condition = list(
    total_count = if ("total_count" %in% names(data_subset))
      stats::median(data_subset$total_count, na.rm = TRUE) else NULL,
    diepte = if ("diepte" %in% names(data_subset)) {
      lv <- levels(droplevels(data_subset$diepte))
      if (any(grepl("0-10", lv))) "0-10" else lv[1]
    } else NULL
  ),
  type = tp
)

library(ggplot2)
ggplot(pred, aes(x, predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15) +
  geom_line() +
  facet_wrap(~ group, scales = "free_y") +
  labs(x = "Carbon density", y = "Predicted value") +
  theme_bw()

```


Per–land-use effect sizes (IRR) for c_density

IRR per IQR, work on the link (latent) scale and exponentiate

```{r}
library(emmeans)
iqr_cd <- IQR(data_subset$c_density, na.rm = TRUE)

# Get slopes on the link (log) scale
tr <- summary(
  emtrends(mod, ~ landgebruik, var = "c_density", mode = "latent", infer = TRUE)
) # columns include: landgebruik, c_density.trend, SE, df (Inf), p.value

zcrit <- qnorm(0.975)

irr_tbl <- as.data.frame(tr) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = c_density.trend,
    SE    = SE,
    IRR   = exp(beta * iqr_cd),
    LCL   = exp((beta - zcrit * SE) * iqr_cd),
    UCL   = exp((beta + zcrit * SE) * iqr_cd),
    p.value
  )

irr_tbl


```

keep delta inside emtrends(), do this (still on the latent scale) and then exponentiate the trend directly:

```{r}
tr_delta <- summary(
  emtrends(mod, ~ landgebruik, var = "c_density",
           mode = "latent", delta = iqr_cd, infer = TRUE)
)
irr_tbl <- as.data.frame(tr_delta) |>
  dplyr::transmute(
    landgebruik,
    increment = paste0("per IQR (", round(iqr_cd, 2), ")"),
    beta  = c_density.trend,  # already scaled by delta
    SE    = SE,               # SE is scaled too
    IRR   = exp(beta),
    LCL   = exp(beta - qnorm(0.975) * SE),
    UCL   = exp(beta + qnorm(0.975) * SE),
    p.value
  )

```


Center/scale c_density (optional) to stabilize estimation and make the intercept interpretable? Not done yet

```{r eval=FALSE}
data_subset$c_density_c <- scale(data_subset$c_density, center = TRUE, scale = TRUE)
# then use c_density_c instead of c_density in both above formulas

```

