### `r stringr::str_to_sentence("{{group}} - {{primerset}} - {{unit}}")`

```{r}
group <- "{{group}}"
primerset <- "{{primerset}}"
unit <- "{{unit}}"
season_run <- "{{season_run}}"

```

Seasonal filtering inside the child

```{r}
# Seasonal filtering inside the child
months_map <- list(
  allseasons     = NULL,
  winterhalfjaar = c(11, 12, 1, 2, 3, 4),
  winter         = c(12, 1, 2, 3, 4)
)

data_subset <- combined %>%
  {
    sel <- months_map[[season_run]]
    if (!is.null(sel) && "sampling_month" %in% names(.)) {
      dplyr::filter(., sampling_month %in% sel)
    } else {
      .
    }
  } %>%
  dplyr::filter(
    group     == "{{group}}",
    primerset == "{{primerset}}",
    unit      == "{{unit}}"
  ) %>%
  droplevels()
```

```{r}
cat("season_run =", season_run,
    "| rows =", nrow(data_subset),
    "| months =",
    if ("sampling_month" %in% names(data_subset))
      paste(sort(unique(data_subset$sampling_month)), collapse = ",")
    else "NA",
    "\n")

```

basic statistics

```{r}
data_subset %>%
  group_by(landgebruik, diepte) %>%
  summarize(n_shannon_not_missing = sum(!is.na(shannon))) %>%
  knitr::kable()
```


```{r}
# remove NAs
data_subset <- data_subset %>%
  filter(!is.na(shannon))

diepte_ok <- length(unique(data_subset$diepte)) > 1
```

Fitting models

```{r m-shannon-{{group}}-{{primerset}}-{{unit}}}
if (diepte_ok) {
  form_shannon <- formula(
  shannon ~
    log(total_count + 1)
    + landgebruik
    + diepte
    + landgebruik:diepte
    + (1 | cmon_plot_id)
  )
} else {
  form_shannon <- formula(
  shannon ~
    log(total_count + 1)
    + landgebruik
  + (1 | cmon_plot_id)
  )
}

zeroes <- min(data_subset$shannon, na.rm = TRUE) == 0

if (!zeroes) {
  cat("Fitting Gamma model")

  m_shannon <- glmmTMB(
    formula = form_shannon,
    data = data_subset,
    family = Gamma(link = "log")
  )
} else {
  cat(
    "Zeroes detected (only one taxum observed):
    Fitting zero-inflated Gamma model"
  )

  m_shannon <- glmmTMB(
    formula = form_shannon,
    ziformula = ~1,
    data = data_subset,
    family = ziGamma(link = "log")
  )
}
```

Performance check selected model

```{r m-shannon-checks-{{group}}-{{primerset}}-{{unit}}}
performance::check_model(m_shannon)
```

Summary selected model

```{r m-shannon-summary-{{group}}-{{primerset}}-{{unit}}}
summary(m_shannon) |> print(digits = 2)
```

Capture model family

```{r}
# ---- Shannon: capture model family (NEW) ----
get_family_name <- function(model) {
  tryCatch({
    fam <- stats::family(model)
    if (is.list(fam) && !is.null(fam$family)) as.character(fam$family)
    else if (inherits(fam, "family") && !is.null(fam$family)) fam$family
    else if (is.character(fam)) fam
    else NA_character_
  }, error = function(e) NA_character_)
}

model_family_shannon <- get_family_name(m_shannon)

append_model_family_shannon(
  group        = group,
  primerset    = primerset,
  season_run   = season_run,
  model_family = model_family_shannon
)

```

Anova selected model

```{r m-shannon-anova-{{group}}-{{primerset}}-{{unit}}}
car::Anova(m_shannon) |> print(digits = 2)
```

Estimate contrasts between land-use categories per depth

```{r m-shannon-emmeans-contrasts-per-depth-{{group}}-{{primerset}}-{{unit}}}
# Estimate marginal means per land use per depth
emm_shannon_landuse_by_depth <- emmeans(
  m_shannon,
  specs = "landgebruik",
  by = if (diepte_ok) "diepte" else NULL,
  data = data_subset,
  type = "response"
)

# Pairwise contrasts (auto adjustment by emmeans; keep CIs & p-values)
contrasts_shannon_landuse_response <- contrast(
  emm_shannon_landuse_by_depth,
  method = "pairwise",
  infer = c(TRUE, TRUE)
) %>%
  as.data.frame()

# Save to CSV
write_csv(
  contrasts_shannon_landuse_response,
  file = file.path(output_dir, glue("contrasts_landuse_shannon_{group}_{primerset}_{unit}.csv"))
)

# Print
print(contrasts_shannon_landuse_response)

```

Accumulator contrasts between land-use categories per depth

```{r}
# Summary table
contrasts_tbl_sh <- contrasts_shannon_landuse_response %>%
  as_tibble() %>%
  dplyr::select(
    dplyr::any_of(c(
      "diepte",        # only present if diepte_ok == TRUE
      "contrast",
      "estimate",      # default column name
      "ratio",         # appears instead of 'estimate' when type = "response"
      "SE",
      "z.ratio",       # or t.ratio, one of them will always exist with infer=TRUE
      "p.value",
      "asymp.LCL",
      "asymp.UCL"
    ))
  )

# Sort
if (diepte_ok) {
  contrasts_tbl_sh <- contrasts_tbl_sh %>% arrange(diepte, p.value)
} else {
  contrasts_tbl_sh <- contrasts_tbl_sh %>% arrange(p.value)
}

# Save
write_csv(
  contrasts_tbl_sh,
  file = file.path(output_dir, glue("contrasts_landuse_table_response_{group}_{primerset}_{unit}.csv"))
)

if (!"z.ratio" %in% names(contrasts_shannon_landuse_response) && "t.ratio" %in% names(contrasts_shannon_landuse_response)) {
  contrasts_shannon_landuse_response$z.ratio <- contrasts_shannon_landuse_response$t.ratio
}

if (!"asymp.LCL" %in% names(contrasts_shannon_landuse_response) && "lower.CL" %in% names(contrasts_shannon_landuse_response)) {
  contrasts_shannon_landuse_response$asymp.LCL <- contrasts_shannon_landuse_response$lower.CL
  contrasts_shannon_landuse_response$asymp.UCL <- contrasts_shannon_landuse_response$upper.CL
}


# Build a tidy table with standard columns
contrasts_tbl_shannon <- contrasts_shannon_landuse_response %>%
  # Ensure 'diepte' column exists even if diepte_ok == FALSE
  { if (!"diepte" %in% names(.)) mutate(., diepte = NA_character_) else . } %>%
  # Split "Akker / Natuurgrasland" -> two columns
  mutate(contrast = as.character(contrast)) %>%
  tidyr::separate(contrast, into = c("landgebruik_1", "landgebruik_2"),
                  sep = " / ", remove = TRUE, fill = "right") %>%
  mutate(
    effect      = .data$ratio,
    effect_pct  = 100 * (effect - 1),  # percent difference (numerator vs denominator)
    lcl         = .data$asymp.LCL,
    ucl         = .data$asymp.UCL,
    stat        = .data$z.ratio,
    stat_name   = dplyr::case_when(!is.na(.data$z.ratio) ~ "z",
                                   TRUE ~ NA_character_)
  ) %>%
  # Tag with run identifiers and order columns
  mutate(
    group      = group,
    primerset  = primerset,
    season_run = season_run
  ) %>%
  relocate(group, primerset, season_run, diepte, landgebruik_1, landgebruik_2,
           effect, effect_pct, SE, lcl, ucl, stat_name, stat, p.value)

# Append to the parent accumulator
append_summary_contrasts_shannon(contrasts_tbl_shannon)

# Print
contrasts_tbl_sh
```


Marginal effects average predictions (double check marginal means but different method)

```{r m-shannon-preds-values-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
predicted_values <- marginaleffects::avg_predictions(
  m_shannon,
  variables = c("landgebruik", "diepte"),
  newdata = data_subset,
  type = "response"
)

# save to CSV
write_csv(
  predicted_values,
  file = file.path(output_dir, glue("predicted_values_shannon_{group}_{primerset}_{unit}.csv"))
)

# View
print(predicted_values)
```

Plot predictions total count

```{r m-shannon-preds-total-count-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_shannon,
  condition = c("total_count"),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Shannon index") +
  scale_x_log10()
```

Plot predictions design variables

```{r m-shannon-preds-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_shannon,
  condition = c("landgebruik", "diepte"[diepte_ok]),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Shannon index")
```


