
### `r stringr::str_to_sentence("{{group}} - {{primerset}} - {{unit}}")`

```{r}
group <- "{{group}}"
primerset <- "{{primerset}}"
unit <- "{{unit}}"
season_run <- "{{season_run}}"

```

Seasonal filtering inside the child

```{r}
# Seasonal filtering inside the child
months_map <- list(
  allseasons     = NULL,
  winterhalfjaar = c(11, 12, 1, 2, 3, 4),
  winter         = c(12, 1, 2, 3, 4)
)

data_subset <- combined %>%
  {
    sel <- months_map[[season_run]]
    if (!is.null(sel) && "sampling_month" %in% names(.)) {
      dplyr::filter(., sampling_month %in% sel)
    } else {
      .
    }
  } %>%
  dplyr::filter(
    group     == "{{group}}",
    primerset == "{{primerset}}",
    unit      == "{{unit}}"
  ) %>%
  droplevels()
```

```{r}
cat("season_run =", season_run,
    "| rows =", nrow(data_subset),
    "| months =",
    if ("sampling_month" %in% names(data_subset))
      paste(sort(unique(data_subset$sampling_month)), collapse = ",")
    else "NA",
    "\n")

```

basic statistics

```{r}
data_subset %>%
  group_by(landgebruik, diepte) %>%
  summarize(n_shannon_not_missing = sum(!is.na(shannon))) %>%
  knitr::kable()
```


```{r}
# remove NAs
data_subset <- data_subset %>%
  filter(!is.na(shannon))

diepte_ok <- length(unique(data_subset$diepte)) > 1
```

#### Basismodel

Fitting models

```{r m-shannon-{{group}}-{{primerset}}-{{unit}}}
if (diepte_ok) {
  form_shannon <- formula(
  shannon ~
    log(total_count + 1)
    + landgebruik
    + diepte
    + landgebruik:diepte
    + (1 | cmon_plot_id)
  )
} else {
  form_shannon <- formula(
  shannon ~
    log(total_count + 1)
    + landgebruik
  + (1 | cmon_plot_id)
  )
}

zeroes <- min(data_subset$shannon, na.rm = TRUE) == 0

if (!zeroes) {
  cat("Fitting Gamma model")

  m_shannon <- glmmTMB(
    formula = form_shannon,
    data = data_subset,
    family = Gamma(link = "log")
  )
} else {
  cat(
    "Zeroes detected (only one taxum observed):
    Fitting zero-inflated Gamma model"
  )

  m_shannon <- glmmTMB(
    formula = form_shannon,
    ziformula = ~1,
    data = data_subset,
    family = ziGamma(link = "log")
  )
}
```

Performance check selected model

```{r m-shannon-checks-{{group}}-{{primerset}}-{{unit}}}
performance::check_model(m_shannon)
```

Summary selected model

```{r m-shannon-summary-{{group}}-{{primerset}}-{{unit}}}
summary(m_shannon) |> print(digits = 2)
```

Capture model family

```{r}
# ---- Shannon: capture model family (NEW) ----
get_family_name <- function(model) {
  tryCatch({
    fam <- stats::family(model)
    if (is.list(fam) && !is.null(fam$family)) as.character(fam$family)
    else if (inherits(fam, "family") && !is.null(fam$family)) fam$family
    else if (is.character(fam)) fam
    else NA_character_
  }, error = function(e) NA_character_)
}

model_family_shannon <- get_family_name(m_shannon)

append_model_family_shannon(
  group        = group,
  primerset    = primerset,
  season_run   = season_run,
  model_family = model_family_shannon
)

```

Anova selected model

```{r m-shannon-anova-{{group}}-{{primerset}}-{{unit}}}
car::Anova(m_shannon) |> print(digits = 2)
```

Estimate contrasts between land-use categories per depth

```{r m-shannon-emmeans-contrasts-per-depth-{{group}}-{{primerset}}-{{unit}}}
# --- EMMs per land use (and depth if applicable) on the response (H′) scale
emm_shannon_landuse_by_depth <- emmeans(
  m_shannon,
  specs = "landgebruik",
  by    = if (diepte_ok) "diepte" else NULL,
  data  = data_subset,
  type  = "response",
  component = "response"   # crucial for ziGamma: overall ('global') mean incl. zeros
)

# Put contrasts on the identity scale (H′) so pairwise are differences, not ratios
emm_shannon_resp <- regrid(emm_shannon_landuse_by_depth, transform = "response")

# --- Pairwise *differences* in H′ (ΔH′) with CIs & p-values
pairs_deltaH <- contrast(
  emm_shannon_resp,
  method = "pairwise",
  by     = if (diepte_ok) "diepte" else NULL,
  infer  = c(TRUE, TRUE)
) %>% as.data.frame()


# Ensure 'diepte' column exists for the no-depth case
if (!"diepte" %in% names(pairs_deltaH)) {
  pairs_deltaH$diepte <- NA_character_
}

# Split "A - B" (differences) OR "A / B" (just in case) into two columns
contrasts_tbl_deltaH <- pairs_deltaH %>%
  tibble::as_tibble() %>%
  dplyr::mutate(contrast = as.character(.data$contrast)) %>%
  tidyr::separate(
    contrast,
    into = c("landgebruik_1", "landgebruik_2"),
    sep  = " / | - ",
    remove = TRUE, fill = "right"
  ) %>%
  dplyr::mutate(
    delta_H   = .data$estimate,
    lcl       = .data$asymp.LCL,
    ucl       = .data$asymp.UCL,
    hill_ratio = exp(delta_H),
    hill_LCL   = exp(lcl),
    hill_UCL   = exp(ucl),
    stat       = .data$z.ratio,
    stat_name  = dplyr::case_when(!is.na(.data$z.ratio) ~ "z",
                                  TRUE ~ NA_character_)
  )

print(pairs_deltaH)
```

```{r}
print(contrasts_tbl_deltaH)
```


Accumulator contrasts between land-use categories per depth

```{r}
# --- Per-group EMMs (H′) and Hill numbers, to attach to each side of the pair
emm_tbl <- as.data.frame(emm_shannon_resp) %>%
  { if (!"diepte" %in% names(.)) dplyr::mutate(., diepte = NA_character_) else . } %>%
  dplyr::transmute(
    diepte,
    landgebruik,
    H_mean = emmean,
    H_SE   = SE,
    H_LCL  = asymp.LCL,
    H_UCL  = asymp.UCL,
    Hill_mean = exp(H_mean),
    Hill_LCL  = exp(H_LCL),
    Hill_UCL  = exp(H_UCL)
  )

# Join EMMs for both sides of the contrast
if (isTRUE(diepte_ok)) {
  by1 <- c("diepte" = "diepte", "landgebruik_1" = "landgebruik")
  by2 <- c("diepte" = "diepte", "landgebruik_2" = "landgebruik")
} else {
  by1 <- c("landgebruik_1" = "landgebruik")
  by2 <- c("landgebruik_2" = "landgebruik")
}

contrasts_tbl_shannon <- contrasts_tbl_deltaH %>%
  dplyr::left_join(emm_tbl, by = by1, multiple = "first") %>%
  dplyr::rename(
    H1_mean = H_mean, H1_SE = H_SE, H1_LCL = H_LCL, H1_UCL = H_UCL,
    Hill1_mean = Hill_mean, Hill1_LCL = Hill_LCL, Hill1_UCL = Hill_UCL
  ) %>%
  dplyr::left_join(emm_tbl, by = by2, multiple = "first") %>%
  dplyr::rename(
    H2_mean = H_mean, H2_SE = H_SE, H2_LCL = H_LCL, H2_UCL = H_UCL,
    Hill2_mean = Hill_mean, Hill2_LCL = Hill_LCL, Hill2_UCL = Hill_UCL
  ) %>%
  # Tag with run identifiers and tidy columns
  dplyr::mutate(
    group      = group,
    primerset  = primerset,
    season_run = season_run
  ) %>%
  dplyr::select(
    group, primerset, season_run, diepte,
    landgebruik_1, landgebruik_2,
    # Pairwise contrast results
    delta_H, lcl, ucl, hill_ratio, hill_LCL, hill_UCL,
    SE, stat_name, stat, p.value,
    # EMMs for each side (H′ and Hill)
    H1_mean, H1_SE, H1_LCL, H1_UCL, Hill1_mean, Hill1_LCL, Hill1_UCL,
    H2_mean, H2_SE, H2_LCL, H2_UCL, Hill2_mean, Hill2_LCL, Hill2_UCL
  ) %>%
  { if (diepte_ok) dplyr::arrange(., diepte, p.value) else dplyr::arrange(., p.value) }

# Save per-run table
readr::write_csv(
  contrasts_tbl_shannon,
  file = file.path(output_dir, glue::glue("contrasts_landuse_table_deltaH_hill_withEMMs_{group}_{primerset}_{unit}.csv"))
)

# Append to the parent accumulator (your existing helper) and write summary
append_summary_contrasts_shannon(contrasts_tbl_shannon)


# (Optional) also save raw EMMs for reference
readr::write_csv(
  emm_tbl %>% dplyr::mutate(group = group, primerset = primerset, season_run = season_run),
  file = file.path(output_dir, glue::glue("emmeans_shannon_groups_H_and_Hill_{group}_{primerset}_{unit}.csv"))
)
```


Marginal effects average predictions (double check marginal means but different method)

```{r m-shannon-preds-values-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
predicted_values <- marginaleffects::avg_predictions(
  m_shannon,
  variables = c("landgebruik", "diepte"),
  newdata = data_subset,
  type = "response"
)

# save to CSV
write_csv(
  predicted_values,
  file = file.path(output_dir, glue("predicted_values_shannon_{group}_{primerset}_{unit}.csv"))
)

# View
print(predicted_values)
```

Plot predictions total count

```{r m-shannon-preds-total-count-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_shannon,
  condition = c("total_count"),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Shannon index") +
  scale_x_log10()
```

Plot predictions design variables

```{r m-shannon-preds-landgebruikxdiepte-{{group}}-{{primerset}}-{{unit}}}
marginaleffects::plot_predictions(
  m_shannon,
  condition = c("landgebruik", "diepte"[diepte_ok]),
  re.form = NA,
  vcov = TRUE,
  type = "response"
) +
  labs(y = "Shannon index")
```

#### Met fysicochemische data

Stapsgewijze modelselectie:

```{r}
variables_no_collinearity <- c(
  "total_count",
  "landgebruik",
  "diepte"[diepte_ok],
  "cmon_plot_id",
  "c_density",
  "swc_vol",
  "ph_kcl",
  "textuur_som_klei_zand"
)

nr1 <- nrow(data_subset)
cat(
  paste(
    "De originele dataset heeft",
    nr1,
    "rijen.\n"
  )
)
data_subset <- data_subset %>%
  mutate(log_c_density = log(c_density)) %>%  # log-transform predictor
  { 
    # Guard: non-positive c_density cannot be logged
    if (any(!is.finite(.$log_c_density))) {
      stop("c_density bevat niet-positieve waarden; log() is niet gedefinieerd. Filter/aanpassen vereist.")
    } else .
  } %>%
  filter(complete.cases(across(all_of(c(variables_no_collinearity, "log_c_density"))))) %>%
  droplevels()
nr2 <- nrow(data_subset)

cat(
  paste(
    "Na verwijderen van rijen waarvan er ontbrekende gegevens zijn voor
    één van de covariaten, heeft de dataset",
    nr2,
    "rijen.\n"
  )
)
```

We voegen alle niet collineaire fysicochemische variabelen toe aan het basismodel.
Voor pH gebruiken we een kwadratische effect.
Dit wordt het `full model`.

```{r extend-shannon-model-additional-predictors-{{group}}-{{primerset}}-{{unit}}}
form_shannon_full <- update(
  form_shannon,
  . ~ .
    + swc_vol
    + log_c_density
    + poly(ph_kcl, 2)
    + textuur_som_klei_zand
)

m_shannon_full <- update(
  m_shannon,
  formula = form_shannon_full,
  data = data_subset,
  na.action = na.fail
)
```

We vergelijken de parameterschattingen van het basis en volledig model.

```{r coefficients-se-shannon-{{group}}-{{primerset}}-{{unit}}}
base_model_coefs <- broom.mixed::tidy(m_shannon)
full_model_coefs <- broom.mixed::tidy(m_shannon_full)

coefs_df <- rbind(
  data.frame(Model = "Base", base_model_coefs),
  data.frame(Model = "Full", full_model_coefs)
)
```

```{r plot-coefficient-estimates-shannon-{{group}}-{{primerset}}-{{unit}}}
ggplot(
  coefs_df,
  aes(x = term, y = estimate, fill = Model)
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(
    aes(
      ymin = estimate - std.error,
      ymax = estimate + std.error
    ),
    width = 0.2, position = position_dodge(0.9)
  ) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Coefficient Estimates and Standard Errors",
    x = "Predictor",
    y = "Estimate",
    fill = "Model"
  )
```

```{r save-coefficient-estimates-shannon-{{group}}-{{primerset}}-{{unit}}}
readr::write_csv(
  coefs_df,
  file = file.path(
    output_dir,
    glue::glue("coefs_base_vs_full_shannon-{group}_{primerset}_{unit}.csv")
  )
)
```

Modelselectie met de functie `dredge`.
We beperken het aantal te verkennen parametercombinaties door te forceren dat `total count` `landgebruik` en `diepte` altijd geschat moeten worden ('by design') en maximaal 7 termen in het model zitten.

```{r model-selection-and-comparison-shannon-{{group}}-{{primerset}}-{{unit}}, warning=FALSE}
if (diepte_ok) {
  model_set <- dredge(
    m_shannon_full,
    m.lim = c(NA, 7),
    fixed = ~
      cond(log(total_count + 1))
    + cond(landgebruik)
    + cond(diepte)
    + (1 | cmon_plot_id)
  )
} else {
  model_set <- dredge(
    m_shannon_full,
    m.lim = c(NA, 7),
    fixed = ~
      cond(log(total_count + 1))
    + cond(landgebruik)
    + (1 | cmon_plot_id)
  )
}
```

Per variabele berekenen we de som van modelgewichten over alle modellen en tellen we in hoeveel modellen de variabele voorkwam:

```{r importance-per-predictor-shannon-{{group}}-{{primerset}}-{{unit}}}
importance_df <- sw(model_set)
print(importance_df)
```

Visualisatie van modellen die minder dan 4 AICc verschillen van het model met laagste AICc (rij 1):

```{r}
dd <- subset(model_set, delta < 4)

par(mar = c(3, 5, 6, 4))
plot(dd, labAsExpr = TRUE)
```

Dit is de samenvatting voor het model met de laagste AICc:

```{r}
lowest_aic <- get.models(dd, 1)[[1]]
summary(lowest_aic)
```

```{r m-shannon-anova-incl-fyschem-{{group}}-{{primerset}}-{{unit}}}
car::Anova(lowest_aic) |> print(digits = 2)
```

plot model diagnostics and model predictions

```{r error=TRUE}
p <- performance::check_predictions(lowest_aic)
plot(p) + facet_zoom(xlim = c(0, 10))
p <- performance::check_overdispersion(lowest_aic)
p
p %>% plot()
p <- performance::check_residuals(lowest_aic)
p
p %>% plot()
performance::check_collinearity(lowest_aic)
performance::check_zeroinflation(lowest_aic)
```


```{r}
at <- car::Anova(lowest_aic)
efvars <- rownames(at)[order(-at$Chisq)]
interactingvars <- efvars[grepl(":", efvars)]
interactingvars <- stringr::str_split(interactingvars, ":") %>%
  unlist() %>%
  unique()
efvars <- efvars[!efvars %in% interactingvars]
efvars <- stringr::str_replace_all(
  efvars, "^(\\w+\\()?(.+?)(\\))?$", "\\2"
)
efvars <- ifelse(
  efvars == "diepte:landgebruik", "landgebruik:diepte", efvars
)
efvars <- ifelse(
  efvars == "total_count + 1", "total_count", efvars
)
efvars <- ifelse(
  efvars == "ph_kcl, 2", "ph_kcl", efvars
)

p <- vector(mode = "list", length = length(efvars))
p <- set_names(p, efvars)
for (i in efvars) {
  if (grepl(":", i)) {
    conds <- stringr::str_split_1(i, ":")
    nd <- insight::get_datagrid(
      lowest_aic,
      by = conds,
      length = 50
    )
    pl <- marginaleffects::plot_predictions(
      model = lowest_aic,
      by = conds,
      newdata = nd,
      vcov = TRUE,
      type = "response",
      points = 0.1,
      re.form = NA
    )
  } else {
    nd <- insight::get_datagrid(
      lowest_aic,
      by = i,
      length = 50
    )
    pl <- marginaleffects::plot_predictions(
      lowest_aic,
      by = i,
      newdata = nd,
      vcov = TRUE,
      type = "response",
      points = 0.1,
      re.form = NA
    )
  }
  p[[i]] <- pl
}

p
```

global log_c_density effect. This gives a global (averaged) slope across the different land-use categories
```{r log-c-density-lowest_aic-shannon-{{group}}-{{primerset}}-{{unit}}}

# Slope for log(c_density) on the latent (log) scale
tr_add <- emtrends(lowest_aic, ~ 1, var = "log_c_density", mode = "latent")
tr_add_sum <- summary(tr_add, infer = TRUE)

z <- qnorm(0.975)

coef_tbl <- as_tibble(tr_add_sum) %>%
  transmute(beta = log_c_density.trend, SE = SE, p.value = p.value)

inc_tbl <- tibble(
  increment = c("×e (e-fold)", "×2 (doubling)", "×10 (tenfold)", "×1.2 (+20%)"),  
  delta     = c(1,               log(2),          log(10),         log(1.2)),
  code      = c("e",             "2x",            "10x",           "1.2x")   # NEW      
)

result <- crossing(coef_tbl, inc_tbl) %>%
  mutate(
    IRR = exp(beta * delta),
    LCL = exp((beta - z * SE) * delta),
    UCL = exp((beta + z * SE) * delta)
  )

result
```

```{r}
# Requires: result (3-row tibble), group, primerset, season_run
# Also assumes you've already created 'summary_cdensity_shannon_all' in the parent.
if (!exists("summary_cdensity_shannon_all")) summary_cdensity_shannon_all <- tibble::tibble()

library(dplyr)
library(tidyr)
library(stringr)

.schema <- c(
  "group","primerset","landgebruik","season_run",
  "beta","SE",
  "IRR_e","LCL_e","UCL_e",
  "IRR_2x","LCL_2x","UCL_2x",
  "IRR_1.2x","LCL_1.2x","UCL_1.2x",   # NEW
  "IRR_10x","LCL_10x","UCL_10x",
  "p.value"
)

# Convert the 3 increments (e, 2x, 10x) into one wide row with matching names
global_row <- {
  vals <- result %>%
    summarise(
      beta    = first(beta),
      SE      = first(SE),
      p.value = first(p.value)
    )

  wide <- result %>%
  dplyr::select(code, IRR, LCL, UCL) %>%
  tidyr::pivot_wider(
    names_from  = code,
    values_from = c(IRR, LCL, UCL),
    names_glue  = "{.value}_{code}"
  )

  bind_cols(vals, wide) %>%
    mutate(
      group       = group,
      primerset   = primerset,
      landgebruik = "global",
      season_run  = season_run
    ) %>%
    relocate(group, primerset, landgebruik, season_run)
}

# Guarantee full schema exists
missing <- setdiff(.schema, names(global_row))
if (length(missing)) global_row[missing] <- NA_real_
global_row <- global_row %>% select(all_of(.schema))

# Append
summary_cdensity_shannon_all <- bind_rows(summary_cdensity_shannon_all, global_row)


```

Include the interaction between log(c_density) and landuse (to get landuse specific slopes)

```{r}
lowest_aic_int <- update(
  lowest_aic,
  formula.  = ~ . + landgebruik * log_c_density,
  data      = data_subset,
  na.action = na.fail)

```

summary of the model that includes the interaction between log(c_density) and landuse

```{r}
summary(lowest_aic_int)

```

Does this model fit better or worse?

```{r}
anova(lowest_aic, lowest_aic_int, test = "Chisq")  # LRT for interaction

```

Per–land-use IRRs for fold changes in carbon density (no subsetting)

```{r log-c-density-lowest-aic-int-perlanduse-shannon-{{group}}-{{primerset}}-{{unit}}}
# Slopes per landgebruik w.r.t. log(c_density) on latent scale
tr_lu <- summary(
  emtrends(lowest_aic_int, ~ landgebruik,
           var = "log_c_density", mode = "latent", infer = TRUE)
)

z <- qnorm(0.975)
irr_lu_fc <- as_tibble(tr_lu) %>%
  transmute(
    landgebruik,
    beta = log_c_density.trend,
    SE   = SE,
    # IRRs for fold changes:
    IRR_e   = exp(beta * 1),          LCL_e   = exp((beta - z*SE) * 1),          UCL_e   = exp((beta + z*SE) * 1),
    IRR_2x  = exp(beta * log(2)),     LCL_2x  = exp((beta - z*SE) * log(2)),     UCL_2x  = exp((beta + z*SE) * log(2)),
    IRR_1.2x  = exp(beta * log(1.2)), LCL_1.2x = exp((beta - z*SE) * log(1.2)),  UCL_1.2x = exp((beta + z*SE) * log(1.2)),  # NEW
    IRR_10x = exp(beta * log(10)),    LCL_10x = exp((beta - z*SE) * log(10)),    UCL_10x = exp((beta + z*SE) * log(10)),
    p.value
  )

print(irr_lu_fc, n = Inf, width = Inf)   # <- shows every column/row

```

Append the **per-landuse** effects (`irr_lu_fc`) to `summary_cdensity_shannon_all`

```{r}
# Requires: irr_lu_fc (per landgebruik tibble), group, primerset, season_run

if (!exists("summary_cdensity_shannon_all")) summary_cdensity_shannon_all <- tibble::tibble()

.schema <- c(
  "group","primerset","landgebruik","season_run",
  "beta","SE",
  "IRR_e","LCL_e","UCL_e",
  "IRR_2x","LCL_2x","UCL_2x",
  "IRR_1.2x","LCL_1.2x","UCL_1.2x",   # NEW
  "IRR_10x","LCL_10x","UCL_10x",
  "p.value"
)


lu_rows <- irr_lu_fc %>%
  mutate(
    group       = group,
    primerset   = primerset,
    landgebruik = as.character(landgebruik),
    season_run  = season_run
  ) %>%
  relocate(group, primerset, landgebruik, season_run)

# Guarantee full schema exists
missing <- setdiff(.schema, names(lu_rows))
if (length(missing)) lu_rows[missing] <- NA_real_
lu_rows <- lu_rows %>% select(all_of(.schema))

# Append
summary_cdensity_shannon_all <- bind_rows(summary_cdensity_shannon_all, lu_rows)

```

plots per land-use

```{r}
# --- Pull levels & ranges from the fitted model ---
.mf <- model.frame(lowest_aic_int)  # training frame actually used
stopifnot("landgebruik" %in% names(.mf))  # should be there after you added the interaction
lu_lvls <- levels(.mf$landgebruik)

has_diepte <- "diepte" %in% names(.mf) && is.factor(.mf$diepte)
if (has_diepte) {
  dep_lvls      <- levels(.mf$diepte)
  ref_depth_chr <- if (any(grepl("^0-10", dep_lvls))) "0-10" else dep_lvls[1]
  ref_depth     <- factor(ref_depth_chr, levels = dep_lvls)
}

# Range for log_c_density strictly within training data (avoid extrapolation weirdness)
stopifnot("log_c_density" %in% names(.mf))
rng_log <- range(.mf$log_c_density, na.rm = TRUE)
log_seq <- seq(rng_log[1], rng_log[2], length.out = 80)

# --- Build datagrid from the MODEL (auto-fills all other covariates) ---
nd <- insight::get_datagrid(
  lowest_aic_int,
  by     = c("landgebruik", "log_c_density"),
  length = 80
)

# Force the grid to use *all* landgebruik levels & our log grid
nd <- tidyr::crossing(
  landgebruik   = factor(lu_lvls, levels = lu_lvls),
  log_c_density = log_seq
) %>%
  dplyr::left_join(
    # take a single representative row from get_datagrid per landgebruik to carry other covariates
    nd %>%
      dplyr::group_by(landgebruik) %>%
      dplyr::slice(1) %>%
      dplyr::ungroup() %>%
      dplyr::select(-log_c_density),  # we’ll supply our own sequence
    by = "landgebruik"
  ) %>%
  { if (has_diepte) dplyr::mutate(., diepte = ref_depth) else . } %>%
  dplyr::mutate(c_density = exp(log_c_density))

# --- Predictions with CI; fall back to point preds if CI fails ---
pr <- tryCatch(
  marginaleffects::predictions(
    model   = lowest_aic_int,
    newdata = nd,
    type    = "response",
    vcov    = TRUE,
    re.form = NA
  ),
  error = function(e) {
    est <- stats::predict(lowest_aic_int, newdata = nd, type = "response", re.form = NA)
    dplyr::bind_cols(nd, tibble::tibble(estimate = as.numeric(est),
                                        conf.low = NA_real_, conf.high = NA_real_))
  }
)

has_ci <- "conf.low" %in% names(pr) && any(is.finite(pr$conf.low))

# --- Plot ---
p <- ggplot(pr, aes(c_density, estimate))
if (has_ci) p <- p + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15)
p + geom_line(na.rm = TRUE) +
  facet_wrap(~ landgebruik, scales = "free_y") +
  labs(
    x = "Carbon density",
    y = "Voorspelde shannon diversity index",
    title = "Voorspelde shannon diversity index vs carbon density per landgebruik",
    subtitle = paste0(
      "Bij total_count = ",
      if ("total_count" %in% names(.mf)) round(stats::median(.mf$total_count, na.rm = TRUE)) else "—",
      if (has_diepte) paste0(", diepte = ", as.character(ref_depth)) else ""
    )
  ) +
  theme_bw()

```

alternative: plot on the log scale

```{r}
# Reuse .mf, lu_lvls, has_diepte, ref_depth from above

rng_log <- range(.mf$log_c_density, na.rm = TRUE)
log_seq <- seq(rng_log[1], rng_log[2], length.out = 80)

nd_log <- insight::get_datagrid(
  lowest_aic_int,
  by     = c("landgebruik", "log_c_density"),
  length = 80
)

nd_log <- tidyr::crossing(
  landgebruik   = factor(lu_lvls, levels = lu_lvls),
  log_c_density = log_seq
) %>%
  dplyr::left_join(
    nd_log %>%
      dplyr::group_by(landgebruik) %>%
      dplyr::slice(1) %>%
      dplyr::ungroup() %>%
      dplyr::select(-log_c_density),
    by = "landgebruik"
  ) %>%
  { if (has_diepte) dplyr::mutate(., diepte = ref_depth) else . } %>%
  dplyr::mutate(c_density = exp(log_c_density))

pr_log <- tryCatch(
  marginaleffects::predictions(
    model   = lowest_aic_int,
    newdata = nd_log,
    type    = "response",
    vcov    = TRUE,
    re.form = NA
  ),
  error = function(e) {
    est <- stats::predict(lowest_aic_int, newdata = nd_log, type = "response", re.form = NA)
    dplyr::bind_cols(nd_log, tibble::tibble(estimate = as.numeric(est),
                                            conf.low = NA_real_, conf.high = NA_real_))
  }
)

has_ci_log <- "conf.low" %in% names(pr_log) && any(is.finite(pr_log$conf.low))
breaks_nat  <- pretty(range(nd_log$c_density, na.rm = TRUE), n = 5)

p2 <- ggplot(pr_log, aes(log_c_density, estimate))
if (has_ci_log) p2 <- p2 + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15)
p2 + geom_line(na.rm = TRUE) +
  facet_wrap(~ landgebruik, scales = "free_y") +
  scale_x_continuous(
    name   = "Carbon density",
    breaks = log(breaks_nat),
    labels = signif(breaks_nat, 3)
  ) +
  labs(
    y = "Voorspelde shannon diversity index",
    title = "Voorspelde shannon diversity index vs log(carbon density) per landgebruik",
    subtitle = paste0(
      "Bij total_count = ",
      if ("total_count" %in% names(.mf)) round(stats::median(.mf$total_count, na.rm = TRUE)) else "—",
      if (has_diepte) paste0(", diepte = ", as.character(ref_depth)) else ""
    )
  ) +
  theme_bw()

```


