---
title: "`inseKP` dataset"
author: "Hans Van Calster"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
library(knitr)
library(here)
opts_chunk$set(echo = TRUE, out.width = "100%")
opts_knit$set(root.dir = here::here())

if (!"phyloseq" %in% rownames(installed.packages())) {
  remotes::install_github("joey711/phyloseq")
}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
if (!"tidytacos" %in% rownames(installed.packages())) {
  remotes::install_github("lebeerlab/tidytacos")
}
library(tidytacos)
if (!"sccomp" %in% rownames(installed.packages())) {
  remotes::install_github("mangiolalaboratory/sccomp")
}
library(sccomp)

mbag_folder <- "G:/Gedeelde drives/PRJ_MBAG" # nolint

```

# Read data

```{r}
insekp <- readRDS(
  file.path(
    mbag_folder,
    "4c_bodembiodiversiteit",
    "data", "statistiek", "InseKP", "community_composition", "data",
    "insekp.swarm.idtaxa.all.rds"
  )
)
```

```{r}
tt_insekp <- tidytacos::from_phyloseq(insekp)
```


```{r}
tt_insekp_all <- tidytacos::everything(tt_insekp)

names(tt_insekp_all)

scdata <- tt_insekp_all |>
  filter(
    !Landgebruik_MBAG %in% c("Heide", "Moeras")
  ) |>
  mutate(
    Diepte = factor(
      Diepte,
      levels = c("0-10", "10/30"),
      labels = c("0-10", "10-30")
    ),
    Landgebruik_MBAG = factor(
      Landgebruik_MBAG,
      levels = c(
        "Akker", "Tijdelijk grasland", "Blijvend grasland",
        "Residentieel grasland", "Natuurgrasland"
      )
    )
  ) |>
  select(
    count,
    sample_id,
    taxon_id,
    Cmon_PlotID,
    Diepte,
    Landgebruik_MBAG,
    phylum,
    class,
    order,
    family,
    genus,
    species
  )
```

```{r}
scdata <- scdata |>
  mutate(
    species_group = case_when(
      phylum ==  "Annelida" ~ "Annelida",
      class == "Collembola" ~ "Collembola",
      phylum == "Arthropoda" & class != "Collembola" ~
        "Arthropoda excl collembola",
      TRUE ~ NA
    )
  )

scdata |>
  group_by(phylum, class, species_group) |>
  summarise(
    n_species = n_distinct(species),
    n_observations = n()
  )

```

```{r}
scdata <- scdata |>
  filter(
    !is.na(species_group)
  )
```

```{r}
prevalences <- scdata %>%
  group_by(species_group, phylum, class, order, family, genus, species) %>%
  summarise(
    n_observations = n(),
    n_reads = sum(count),
    .groups = "drop"
  ) %>%
  arrange(species_group, -n_observations, -n_reads)

prevalences %>%
  select(species_group, species, n_observations, n_reads) %>%
  kable()
```

```{r}
prevalences <- prevalences |>
  mutate(
    taxon_resolved = gsub("unclassified_", "", species)
  )
```


```{r}
scdata_annelida <- scdata |>
  filter(
    species_group == "Annelida"
  ) |>
  inner_join(
    prevalences,
    by = join_by(phylum, class, order, family, genus, species, species_group)
  )

scdata_annelida |>
  mutate(count_class = cut(
    count, c(1, 10, 100, 1000, Inf),
    include.lowest = TRUE)) |>
  count(taxon_resolved, count_class) |>
  kable()
```


```{r}
# need to aggregate because different swarms/ASV/OTU can have same
# taxon_resolved within a sample
scdata_annelida <- scdata_annelida |>
  group_by(sample_id, taxon_resolved, Cmon_PlotID, Diepte, Landgebruik_MBAG) |>
  summarise(count = sum(count), .groups = "drop")
```

```{r}
# restricting further for testing only, not needed for serious runs
scdata_annelida <- scdata_annelida |>
  group_by(taxon_resolved) |>
  filter(n() > 30) |>
  ungroup()
```


```{r}
sccomp_dir <- here::here(
    "data", "compositional_analysis", "sccomp_draws_files"
  )
fs::dir_create(sccomp_dir)
gi_file <- file.path(sccomp_dir, ".gitignore")
fs::file_create(
  gi_file
)
writeLines(
  text = c("*", "!.gitignore"),
  con = gi_file
)
m1 <- sccomp_estimate(
  .data = scdata_annelida,
  formula_composition =
    ~ Landgebruik_MBAG + Diepte + Landgebruik_MBAG:Diepte + (1 | Cmon_PlotID),
  formula_variability = ~ 1,
  .sample = sample_id,
  .cell_group = taxon_resolved,
  .abundance = count,
  cores = 4,
  output_directory = sccomp_dir, # might want to change this
  bimodal_mean_variability_association = FALSE,
  percent_false_positive = 5,
  inference_method = "pathfinder", # 'hmc'
  # ... passed to cmdstanr method $sampling (igv hmc)
  verbose = FALSE
)

```

```{r}
head(m1, n = 10)
```

```{r eval=FALSE}
test <- cmdstanr::as_cmdstan_fit(
  file.path(
    sccomp_dir,
    "glm_multi_beta_binomial-202412051111-1-6c1f45.csv")
)

# the model contains 50 x 80 = 4000 posterior samples for 31875 parameters
# (incl transformed parameters)
test$metadata$method # pathfinder
test$metadata$history_size # 100
test$metadata$num_paths # 50
test$metadata$num_draws # 80
test$metadata$stan_variable_sizes$beta # 10 beta's for 25 taxa
test$metadata$stan_variable_sizes$alpha # an intercept for 25 taxa
test$metadata$stan_variable_sizes$random_effect # 211 random intercepts for 24
# etcetera
pars <- lapply(
  test$metadata$stan_variable_sizes, FUN = \(x) matrixStats::product(x)
)
sum(unlist(pars)) # 31875
```


Check Rhat!

- Rhat < 1.01 is OK
- Rhat between 1.01 and 1.1 is worrying
- Rhat > 1.1 bad


```{r}
hist(m1$c_rhat)
```

Check ESS (effective sample size)!

- ESS > 100 is OK
- ESS between 20 and 100 might be enough
- ESS < 20 problematic

```{r}
hist(m1$c_ess_bulk)
hist(m1$c_ess_tail)
```

If problems:

- try `"hmc"` algorithm
- increase number of iterations

Calculate `FDR`:

E.g. which taxa differ significantly from `akker` can be calculated with contrasts:

```{r}
threshold <- 0.1
m1_contrast <- sccomp_test(
  m1,
  contrasts = c(
    a_ng_010 = "Landgebruik_MBAGNatuurgrasland",
    a_ng_1030 = "-`Diepte10-30` + Landgebruik_MBAGNatuurgrasland + `Landgebruik_MBAGNatuurgrasland:Diepte10-30`", # nolint
    a_tg_010 = "`Landgebruik_MBAGTijdelijk grasland`",
    a_tg_1030 = "-`Diepte10-30` + `Landgebruik_MBAGTijdelijk grasland` + `Landgebruik_MBAGTijdelijk grasland:Diepte10-30`", # nolint
    a_bg_010 = "`Landgebruik_MBAGBlijvend grasland`",
    a_bg_1030 = "-`Diepte10-30` + `Landgebruik_MBAGBlijvend grasland` + `Landgebruik_MBAGBlijvend grasland:Diepte10-30`", # nolint
    a_rg_010 = "`Landgebruik_MBAGResidentieel grasland`",
    a_rg_1030 = "-`Diepte10-30` + `Landgebruik_MBAGResidentieel grasland` + `Landgebruik_MBAGResidentieel grasland:Diepte10-30`" # nolint
    ),
  test_composition_above_logit_fold_change = threshold
)

```

These contrasts are in logit (=log-odds) scale and represent log-odds ratios:

$$
\begin{align*}
\log\left(\frac{P_{R_{0-10}}}{1-P_{R_{0-10}}}\right) = \text{log-odds } R_{0-10}
\tag*{Intercept parameter: Akker (R for reference) and 0-10 cm}\\

\text{log odds } T_{0-10} - \text{log odds } R_{0-10} & =
\log\left(\frac{\text{odds } T_{0-10}}{\text{odds } R_{0-10}}\right)
\tag*{Land-use T parameter = difference contrast}
\end{align*}
$$

We can transform the difference contrasts from the log-odds differences to odds-ratios for visualisation and easier interpretation.

A value of 2 means that the odds one of the grassland types for the taxon is double the odds in agricultural fields.
Conversely, an odds-ratio equal to 0.5 means that the odds for the taxon is half of the odds in agricultural fields.
Only taxa that are significant after `FDR` correction are shown.

```{r}
m1_plots <- m1_contrast |>
  filter(
    c_FDR <= 0.05
  ) |>
  mutate(
    Diepte = ifelse(grepl("010", parameter), "0-10", "10-30"),
    Landgebruik = gsub("^a_(.*)_\\d+", "\\1", parameter),
    Landgebruik = factor(
      Landgebruik,
      levels = c("tg", "bg", "rg", "ng"),
      labels = c("Tijdelijk grasland", "Blijvend grasland",
                 "Residentieel grasland", "Natuurgrasland")
    )) |>
  group_by(Landgebruik) |>
  nest() |>
  mutate(
    plot = purrr::map2(
      data, Landgebruik,
      function(x, y) {
        x$taxon_resolved <- reorder(x$taxon_resolved, x$c_effect)
        ggplot(x) +
          geom_pointrange(
            aes(
              x = taxon_resolved, y = c_effect, ymin = c_lower, ymax = c_upper,
              colour  = Diepte),
            position = position_dodge(width = 0.5)
          ) +
          geom_hline(yintercept = 0) +
          geom_hline(yintercept = threshold, alpha = 0.2) +
          geom_hline(yintercept = -threshold, alpha = 0.2) +
          scale_y_continuous(
            breaks = log(
              c(1 / 50, 1 / 20, 1 / 10, 1 / 5, 1 / 2, 1, 2, 5, 10, 20, 50)),
            labels = \(x) format(exp(x), drop0trailing = TRUE)
          ) +
          coord_flip() +
          labs(
            title = y, y = "Odds-ratio"
          )
        }
    )
  )

patchwork::wrap_plots(
  m1_plots$plot,
  guides = "collect",
  axis_titles = "collect"
  ) &
  patchwork::plot_annotation(title = "Verschillen met akkers")
```


Interpretation aid (the code below does not work; not sure how to fix):

```{r error = TRUE}
m1 |>
  sccomp::sccomp_proportional_fold_change(
    formula_composition = ~ 0 + Landgebruik_MBAG,
    from = "Akker",
    to = "Natuurgrasland"
  )
```

