---
title: "testing differential abundance analysis using R package `rademu`"
author: "Hans Van Calster"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: sentence
---


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
library(here)
opts_chunk$set(echo = TRUE, error = TRUE, out.width = "100%")
opts_knit$set(root.dir = here::here())
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
mbag_bodem_folder <- "G:/Gedeelde drives/PRJ_MBAG/4c_bodembiodiversiteit" # nolint
library(radEmu)
library(phyloseq)
```

# Inlezen data

```{r inlezen}
metadata <- readr::read_csv(
  file.path(
    mbag_bodem_folder,
    "data", 
    "Stratificatie_MBAG_plots",
    "MBAG_stratfile_v2_cleaned_13.csv")
) %>%
  janitor::clean_names() %>%
  rename(
    ph_kcl = p_h_k_cl,
    swc_grav = sw_cgrav,
    swc_vol = sw_cvol,
    cn_stockbased = c_n_stockbased,
    c_density = cdensity,
    n_density = ndensity
  ) %>%
  mutate(
    landgebruik = factor(
      landgebruik_mbag,
      levels = c(
        "Akker", "Tijdelijk grasland", "Blijvend grasland",
        "Residentieel grasland", "Natuurgrasland", "Heide", "Moeras")
    ),
    diepte = gsub("_|/", "-", diepte) |> factor()
  )
load(
  file.path(
    mbag_bodem_folder,
    "data", "statistiek", "Annelida", "phyloseq",
    "physeq_Olig01_Annelida_species.Rdata"
    )
  )
physeq_Olig01_Annelida_species <- physeq_Olig01_Annelida_species |> 
  phyloseq::subset_samples(
    !Landgebruik_MBAG %in% c("Moeras", "Heide")
  )
```



```{r}
sample_data(physeq_Olig01_Annelida_species)$Landgebruik_MBAG <- factor(
  sample_data(physeq_Olig01_Annelida_species)$Landgebruik_MBAG,
  levels = c(
    "Akker", "Tijdelijk grasland", "Blijvend grasland",
    "Residentieel grasland", "Natuurgrasland")
)

```

>Next, we want to confirm that all samples have at least one non-zero count across the categories we’ve chosen and that all categories have at least one non-zero count across the samples we’ve chosen.

```{r}
sum(rowSums(otu_table(physeq_Olig01_Annelida_species)) == 0)
sum(colSums(otu_table(physeq_Olig01_Annelida_species)) == 0)
```

```{r}
# reduce data size a bit, because emuFit on full dataset takes a long time
# here aggregating to genus
physeq_Olig01_Annelida_genus <- physeq_Olig01_Annelida_species |>
  phyloseq::tax_glom(taxrank = "genus")
```



```{r}
m_fit <- emuFit(
  formula = ~ Landgebruik_MBAG + Diepte + Landgebruik_MBAG:Diepte, 
  Y = physeq_Olig01_Annelida_genus,
  cluster = sample_data(physeq_Olig01_Annelida_genus)$Cmon_PlotID,
  run_score_tests = FALSE) 

m_fit$estimation_converged
```

Visualise:

```{r}
coeftab <- as_tibble(m_fit$coef)

genustab <- phyloseq::tax_table(physeq_Olig01_Annelida_genus)

genustbl <- as_tibble(
  genustab@.Data
) |> 
  mutate(
    category = dimnames(genustab)[[1]]) |> 
  rowwise() |> 
  mutate(
    across(
      all_of(c("phylum", "class", "order", "family", "genus")),
      \(x) !grepl("_otu", x),
      .names = "{.col}_"
    ),
    index = sum(c_across(phylum_:genus_)),
    resolved_rank = c("phylum", "class", "order", "family", "genus")[index]
  ) |> 
  ungroup() |> 
  dplyr::select(phylum:category, resolved_rank) |> 
  mutate(
    resolved_name = case_when(
      resolved_rank == "genus" ~ genus,
      resolved_rank == "family" ~ family,
      resolved_rank == "order" ~ order,
      resolved_rank == "class" ~ class,
      TRUE ~ phylum
    ),
    resolved_higher = case_when(
      resolved_rank == "genus" ~ family,
      resolved_rank == "family" ~ family,
      resolved_rank == "order" ~ order,
      resolved_rank == "class" ~ class,
      TRUE ~ phylum
    )
  )

coeftab |>
  inner_join(genustbl) |> 
  mutate(
    genus = factor(genus, levels = unique(genus[order(resolved_higher)]))
  ) |> 
  ggplot() +
  geom_pointrange(
    aes(
      x = genus,
      y = estimate, ymin = lower, ymax = upper,
      colour = resolved_higher)
  ) + 
  facet_wrap(~covariate)

```

Identify taxa for which you want robust score tests:

For instance, which taxa at depth 0 - 10 are more likely to occur or not occur in Natuurgrasland compared to the reference category Akker:

The following code takes a long time to run.
But it can be run in parallel: https://statdivlab.github.io/radEmu/articles/parallel_radEmu.html



```{r}
coefselection <- coeftab |> 
  filter(
    covariate == "Landgebruik_MBAGNatuurgrasland",
    sign(lower) == sign(upper)
  )


covariate_to_test <- which(
  rownames(m_fit$B) == "Landgebruik_MBAGNatuurgrasland")
taxa_to_test <- which(
  coefselection$category %in% colnames(m_fit$B)
  )
m_refit <- emuFit(
  formula = ~ Landgebruik_MBAG + Diepte + Landgebruik_MBAG:Diepte, 
  Y = physeq_Olig01_Annelida_genus,
  test_kj = data.frame(
    k = covariate_to_test, 
    j = taxa_to_test),
  fitted_model = m_fit,
  refit = FALSE,
  run_score_tests = TRUE
)

m_refit$coef$pval[taxa_to_test]
```

Visualise only the significant (p-value <= 0.05) taxa according to the score test:

```{r}
as_tibble(m_refit$coef) |>
  inner_join(genustbl) |>
  filter(
    covariate == "Landgebruik_MBAGNatuurgrasland",
    categorie %in% coefselection$category,
    pval <= 0.05
  ) %>%
  mutate(
    genus = factor(genus, levels = unique(genus[order(resolved_higher)]))
  ) |> 
  ggplot() +
  geom_pointrange(
    aes(
      x = genus,
      y = estimate, ymin = lower, ymax = upper,
      colour = resolved_higher)
  ) +
  labs(
    title = "Natuurgrasland vs Akker (diepte 0-10)"
  )
```





